---
title: "Project in Statistics - Eduardo Soares"
author: "Eduardo Soares"
date: "2026-05-17"
output: html_document
---

#packages used
```{r}
library(dplyr)
library(zoo)
library(ggplot2)
library(expm)
library(patchwork)
```


#Data loading and its preprocessing
```{r}
#Loading the data
excel_url <- "https://www.iceandclimate.nbi.ku.dk/data/GICC05modelext_GRIP_and_GISP2_and_resampled_data_series_Seierstad_et_al._2014_version_10Dec2014-2.xlsx"
res <- httr::GET(excel_url)
excel_file <- tempfile(fileext = ".xlsx")
writeBin(httr::content(res, as = "raw"), excel_file)
Ca2data <- readxl::read_excel(excel_file, sheet = 3,
                              range = "A51:I12278") %>%
  dplyr::slice(-1) %>%
  dplyr::select(1, ncol(.) - 1, ncol(.)) %>%
  dplyr::mutate_all(as.numeric) %>%
  dplyr::rename(age = 1, d18O = 2, Ca2 = 3)
unlink(excel_file)
data_file <- "Ca2data.csv"
readr::write_csv(Ca2data, data_file)

#maybe the Ca2 variable has a really skewed distribution
hist(Ca2data$Ca2, labels = FALSE)
#so like this is maybe better
hist(log(Ca2data$Ca2)) 

Ca2_final <- Ca2data %>%
  select(-d18O) %>%
  filter(age >= 30000, age <= 80000) %>%
  mutate(Ca2 = na.approx(Ca2)) %>%
  group_by(age) %>%
  summarise(Ca2 = mean(Ca2), .groups = "drop") %>%
  mutate(Ca2 = -log(Ca2)) %>%
  mutate(X = Ca2 - mean(Ca2)) %>%
  select(-Ca2) %>%
  mutate(age = age/1000) %>%
  mutate(V = (lead(X) - X)/0.02)

Ca2_final <- Ca2_final %>%
  filter(!is.na(V))


ggplot(Ca2_final, aes(x = age, y = X)) +
  geom_line() +
  scale_x_reverse() +
  labs(
    x = "Age before present (ky)",
    y = "Centered -log(Ca2+)",
    title = ""
  )

Ca2 <- Ca2_final$X
hCa <- 0.02
```


#COMPLETE DATA LIKELIHOOD - CONSTANT NOISE

#Auxiliary functions for the likelihood
```{r}
safe_logdet <- function(M) {
  out <- determinant(M, logarithm = TRUE)
  if (out$sign <= 0) return(Inf)
  as.numeric(out$modulus)
}

quad_form <- function(r, Omega_inv) {
  as.numeric(t(r) %*% Omega_inv %*% r)
}

omega_h <- function(A, Q, h) {
  d <- nrow(A)
  
  K <- kronecker(diag(d), A) + kronecker(A, diag(d))
  I_big <- diag(d * d)
  
  vecQ <- as.vector(Q)
  
  vecOmega <- solve(
    K,
    (expm(K * h) - I_big) %*% vecQ
  )
  
  Omega <- matrix(vecOmega, nrow = d, ncol = d)
  
  0.5 * (Omega + t(Omega))
}

compute_bx <- function(theta, x_data, branch = "+") {
  a <- theta["a"]
  b <- theta["b"]
  
  Ex <- mean(x_data)
  Vx <- var(x_data)
  
  disc <- Vx + (Ex + b / (3 * a))^2
  
  if (branch == "+") {
    -b / (3 * a) + sqrt(disc)
  } else {
    -b / (3 * a) - sqrt(disc)
  }
}


make_A_matrix <- function(theta, x_data, branch = "+") {
  eta <- theta["eta"]
  a   <- theta["a"]
  b   <- theta["b"]
  c   <- theta["c"]
  
  Ex  <- mean(x_data)
  Ex2 <- mean(x_data^2)
  
  lambda <- 3 * a * Ex2 + 2 * b * Ex + c
  bx <- compute_bx(theta, x_data, branch = branch)
  
  A <- matrix(
    c(0,       1,
      lambda, -eta),
    nrow = 2,
    byrow = TRUE
  )
  
  b_vec <- c(bx, 0)
  
  list(
    A = A,
    b_vec = b_vec,
    lambda = lambda,
    bx = bx
  )
}


N_fun <- function(state, theta, lambda, bx) {
  x <- state[1]
  
  a <- theta["a"]
  b <- theta["b"]
  c <- theta["c"]
  d <- theta["d"]
  
  n2 <- a * x^3 + b * x^2 + c * x + d - lambda * (x - bx)
  
  c(0, n2)
}


f_flow <- function(state, t, theta, lambda, bx) {
  state + t * N_fun(state, theta, lambda, bx)
}

f_flow_inv <- function(state, t, theta, lambda, bx) {
  state - t * N_fun(state, theta, lambda, bx)
}

logdet_D_f_inv <- function(state, t, theta, lambda, bx) {
  0
}

mu_h <- function(z, A, b_vec, h) {
  as.numeric(expm(A * h) %*% (z - b_vec) + b_vec)
}
```


#Likelihood implementation
```{r}
ss_negloglik_constSigma_fast <- function(
    par,
    data,
    h,
    branch = "+"
) {
  
  theta <- par[c("eta", "a", "b", "c", "d")]
  gamma <- par["gamma"]
  
  Z <- as.matrix(data)
  colnames(Z) <- c("x", "v")
  
  x_data <- Z[, "x"]
  
  lin <- make_A_matrix(theta, x_data, branch = branch)
  
  A      <- lin$A
  b_vec  <- lin$b_vec
  lambda <- lin$lambda
  bx     <- lin$bx
  
  Sigma <- matrix(c(0, gamma), nrow = 2)
  Q <- Sigma %*% t(Sigma)
  
  Omega <- omega_h(A, Q, h)
  logdet_Omega <- safe_logdet(Omega)
  
  if (!is.finite(logdet_Omega)) return(Inf)
  
  Omega_inv <- solve(Omega)
  
  n <- nrow(Z)
  
  Zk  <- Z[1:(n - 1), , drop = FALSE]
  Zkp <- Z[2:n, , drop = FALSE]
  
  xk  <- Zk[, "x"]
  vk  <- Zk[, "v"]
  xkp <- Zkp[, "x"]
  vkp <- Zkp[, "v"]
  
  a <- theta["a"]
  b <- theta["b"]
  c <- theta["c"]
  d <- theta["d"]
  
  N2_k <- a * xk^3 + b * xk^2 + c * xk + d - lambda * (xk - bx)
  N2_kp <- a * xkp^3 + b * xkp^2 + c * xkp + d - lambda * (xkp - bx)
  
  z_start <- cbind(
    xk,
    vk + (h / 2) * N2_k
  )
  
  y <- cbind(
    xkp,
    vkp - (h / 2) * N2_kp
  )
  
  Phi <- expm(A * h)
  
  m <- t(Phi %*% t(z_start - matrix(b_vec, nrow = n - 1, ncol = 2, byrow = TRUE))) +
    matrix(b_vec, nrow = n - 1, ncol = 2, byrow = TRUE)
  
  R <- y - m
  
  qf <- rowSums((R %*% Omega_inv) * R)
  
  sum(qf + logdet_Omega)
}
```


#Complete data simulation and subsampling
```{r}
EM_sim_fine <- function(N_fine, h_sim, theta, gamma, start = c(0, 0)) {
  X <- numeric(N_fine)
  V <- numeric(N_fine)
  
  X[1] <- start[1]
  V[1] <- start[2]
  
  for (i in 2:N_fine) {
    eps <- rnorm(1)
    
    x_prev <- X[i - 1]
    v_prev <- V[i - 1]
    
    drift_v <- -theta["eta"] * v_prev +
      theta["a"] * x_prev^3 +
      theta["b"] * x_prev^2 +
      theta["c"] * x_prev +
      theta["d"]
    
    X[i] <- x_prev + h_sim * v_prev
    
    V[i] <- v_prev +
      h_sim * drift_v +
      gamma * sqrt(h_sim) * eps
  }
  
  data.frame(x = X, v = V)
}

simulate_subsampled_data <- function(
    N_obs,
    h_obs,
    h_sim,
    theta,
    gamma,
    start = c(0, 0)
) {
  
  ratio <- h_obs / h_sim
  skip <- round(ratio)
  
  N_fine <- (N_obs - 1) * skip + 1
  
  fine_data <- EM_sim_fine(
    N_fine = N_fine,
    h_sim = h_sim,
    theta = theta,
    gamma = gamma,
    start = start
  )
  
  observed_data <- fine_data[seq(1, N_fine, by = skip), ]
  
  rownames(observed_data) <- NULL
  
  observed_data
}
```


#Chosen true parameters, definition of h, h_sim, N, and data simulation
```{r}
theta_true <- c(
  eta = 30,
  a   = -125,
  b   = 40,
  c   = 150,
  d   = -20
)

gamma_true <- sqrt(9000)

h_sim <- 0.0001
h <- 0.01
N_obs <- 5000  


set.seed(52)

dat <- simulate_subsampled_data(
  N_obs = 5000,
  h_obs = h,
  h_sim = h_sim,
  theta = theta_true,
  gamma = gamma_true
)

nrow(dat)

#Sanity check to see if X was well generated
head(dat)
dim(dat)
range(dat$x)
summary(dat$x)
plot(dat$x, type = "l", main = "Simulated X", ylab = "X")
```

#Testing the likelihood once before optimizing it
```{r}
theta_test <- c(
  eta = 50,
  a = -200,
  b = 10,
  c = 100,
  d = 10,
  gamma = sqrt(1000)
)

true_par <- c(
  eta   = 30,
  a     = -125,
  b     = 40,
  c     = 150,
  d     = -20,
  gamma = sqrt(9000)
)

ss_negloglik_constSigma_fast(
  par = true_par,
  data = dat,
  h = h,
  branch = "+"
)

```


#Now the actual optimization
```{r}
eps <- 1e-5

lower_bounds <- c(
  eta   = eps,
  a     = -Inf,
  b     = -Inf,
  c     = -Inf,
  d     = -Inf,
  gamma = eps
)

upper_bounds <- c(
  eta   = Inf,
  a     = -eps,
  b     = Inf,
  c     = Inf,
  d     = Inf,
  gamma = Inf
)


#we can try to optimize it first with only one starting value:
fit <- optim(
  par = theta_test,
  fn = ss_negloglik_constSigma_fast,
  data = dat,
  h = h,
  branch = "+",
  method = "L-BFGS-B",
  lower = lower_bounds,
  upper = upper_bounds,
  control = list(maxit = 1000)
)

#checking the results:
par_hat <- fit$par
fit$value


x_grid <- seq(
  from = min(dat$x, na.rm = TRUE),
  to   = max(dat$x, na.rm = TRUE),
  length.out = 300
)

#Checking if the estimated potential function matches the true potential function
pot_true <- -unname(true_par["a"]) / 4 * x_grid^4 -
  unname(true_par["b"]) / 3 * x_grid^3 -
  unname(true_par["c"]) / 2 * x_grid^2 -
  unname(true_par["d"]) * x_grid

pot_est <- -unname(par_hat["a"]) / 4 * x_grid^4 -
  unname(par_hat["b"]) / 3 * x_grid^3 -
  unname(par_hat["c"]) / 2 * x_grid^2 -
  unname(par_hat["d"]) * x_grid

plot(
  x_grid, pot_true,
  type = "l",
  lwd = 2,
  xlab = "x",
  ylab = "U(x)",
  ylim = range(c(pot_true, pot_est), finite = TRUE)
)

lines(x_grid, pot_est, lwd = 2, lty = 2, col = "red")

legend(
  "topright",
  legend = c("true", "estimated"),
  lty = c(1, 2),
  lwd = 2,
  col = c("black", "red")
)
```


#Function that generates 100 starting points and then fits the optim() with them 
#Also function that generates 500 independent datasets, and plots
```{r}
generate_start_wide <- function() {
  c(
    eta   = runif(1, 1, 150),
    a     = runif(1, -800, -1),
    b     = runif(1, -500, 500),
    c     = runif(1, -300, 600),
    d     = runif(1, -500, 500),
    gamma = runif(1, 5, 400)
  )
}

multi_start_optim <- function(
    n_start,
    data,
    h,
    branch = "+",
    lower_bounds,
    upper_bounds,
    generate_start_fun = generate_start,
    control = list(maxit = 1000)
) {
  
  results <- vector("list", n_start)
  
  for (j in 1:n_start) {
    
    start_j <- generate_start_fun()
    
    fit_j <- optim(
    par = start_j,
    fn = ss_negloglik_constSigma_fast,
    data = data,
    h = h,
    branch = branch,
    method = "L-BFGS-B",
    lower = lower_bounds,
    upper = upper_bounds,
    control = control
  )
    
    results[[j]] <- data.frame(
      start_id = j,
      start_eta   = start_j["eta"],
      start_a     = start_j["a"],
      start_b     = start_j["b"],
      start_c     = start_j["c"],
      start_d     = start_j["d"],
      start_gamma = start_j["gamma"],
      
      eta   = fit_j$par["eta"],
      a     = fit_j$par["a"],
      b     = fit_j$par["b"],
      c     = fit_j$par["c"],
      d     = fit_j$par["d"],
      gamma = fit_j$par["gamma"],
      gamma2 = fit_j$par["gamma"]^2,
      
      value = fit_j$value,
      convergence = fit_j$convergence,
      message = ifelse(is.null(fit_j$message), NA, fit_j$message)
    )
    
    cat("Finished start", j, "with value", fit_j$value, "\n")
  }
  
  do.call(rbind, results)
}

multi_results_wide <- multi_start_optim(
  n_start = 100,
  data = dat,
  h = h,
  branch = "+",
  lower_bounds = lower_bounds,
  upper_bounds = upper_bounds,
  generate_start_fun = generate_start_wide
)

successful <- subset(multi_results_wide, convergence == 0)

par_names <- c("eta", "a", "b", "c", "d", "gamma")

par(mfrow = c(2, 3))

for (p in par_names) {
  hist(
    successful[[p]],
    breaks = 30,
    main = paste("Histogram of", p),
    xlab = p,
    col = "gray",
    border = "white"
  )
  
  abline(v = true_par[p], col = "red", lwd = 2)
}


apply(successful[, c("eta", "a", "b", "c", "d", "gamma")], 2, range)
apply(successful[, c("eta", "a", "b", "c", "d", "gamma")], 2, sd)

range(successful$value)


# Repeated simulation over different datasets

estimate_one_dataset <- function(
    dataset_id,
    N_obs,
    h,
    h_sim,
    theta_true,
    gamma_true,
    theta_start,
    branch = "+",
    maxit = 1000
) {
  
  dat_b <- simulate_subsampled_data(
    N_obs = N_obs,
    h_obs = h,
    h_sim = h_sim,
    theta = theta_true,
    gamma = gamma_true,
    start = c(0, 0)
  )
  
  fit_b <- optim(
    par = theta_start,
    fn = ss_negloglik_constSigma_fast,   # change this if your function has another name
    data = dat_b,
    h = h,
    branch = branch,
    method = "L-BFGS-B",
    lower = lower_bounds,
    upper = upper_bounds,
    control = list(maxit = maxit)
  )
  
  data.frame(
    dataset = dataset_id,
    
    eta   = fit_b$par["eta"],
    a     = fit_b$par["a"],
    b     = fit_b$par["b"],
    c     = fit_b$par["c"],
    d     = fit_b$par["d"],
    gamma = fit_b$par["gamma"],
    
    
    value = fit_b$value,
    convergence = fit_b$convergence,
    message = ifelse(is.null(fit_b$message), NA, fit_b$message)
  )
}

# Run B independent datasets

B <- 500
N_obs <- 5000

theta_start <- true_par

set.seed(123)

sim_results <- vector("list", B)

for (b_id in 1:B) {
  
  sim_results[[b_id]] <- estimate_one_dataset(
    dataset_id = b_id,
    N_obs = N_obs,
    h = h,
    h_sim = h_sim,
    theta_true = theta_true,
    gamma_true = gamma_true,
    theta_start = theta_start,
    branch = "+",
    maxit = 1000
  )
  
  cat("Finished dataset", b_id, "of", B, "\n")
}

sim_results <- do.call(rbind, sim_results) 

table(sim_results$convergence)

sim_successful <- subset(sim_results, convergence == 0)

par_names <- c("eta", "a", "b", "c", "d", "gamma")

estimate_means <- apply(sim_successful[, par_names], 2, mean)
estimate_sds   <- apply(sim_successful[, par_names], 2, sd)
estimate_bias  <- estimate_means - true_par[par_names]

summary_table <- data.frame(
  true_value = true_par[par_names],
  mean_estimate = estimate_means,
  sd_estimate = estimate_sds,
  bias = estimate_bias
)

summary_table

norm_errors <- data.frame(
  eta = (sim_successful$eta - true_par["eta"]) / true_par["eta"],
  a   = (sim_successful$a   - true_par["a"])   / true_par["a"],
  b   = (sim_successful$b   - true_par["b"])   / true_par["b"],
  c   = (sim_successful$c   - true_par["c"])   / true_par["c"],
  d   = (sim_successful$d   - true_par["d"])   / true_par["d"],
  gamma = (sim_successful$gamma - true_par["gamma"]) / true_par["gamma"]
)

for (p in names(norm_errors)) {
  
  dens <- density(norm_errors[[p]])
  
  plot(
    dens,
    main = paste("Normalized error of", p),
    xlab = "",
    lwd = 2
  )
  
  abline(v = 0, col = "red", lwd = 2)
}

```


#Now we turn to partial data likelihood, still with constant noise

#We reuse some auxiliary functions from the complete data likelihood, these are:
#safe_logdet 
#omega_h
#compute_bx
#make_A_matrix
#N_fun
#f_flow
#f_flow_inv
#mu_h
```{r}
logdet_Dv_f_forward <- function(state, t, theta, lambda, bx) {
  0
}


ss_negloglik_partial_R <- function(
    par,
    x_obs,
    h,
    branch = "+",
    return_terms = FALSE
) {
  
  theta <- par[c("eta", "a", "b", "c", "d")]
  gamma <- par["gamma"]
  
  N <- length(x_obs)

  Y <- cbind(
    x = x_obs[1:(N - 1)],
    v = (x_obs[2:N] - x_obs[1:(N - 1)]) / h
  )
  
  x_data <- Y[, "x"]
  
  lin <- make_A_matrix(theta, x_data, branch = branch)
  
  A      <- lin$A
  b_vec  <- lin$b_vec
  lambda <- lin$lambda
  bx     <- lin$bx
  
  
  Sigma <- matrix(
    c(0, gamma),
    nrow = 2,
    ncol = 1
  )
  
  Q <- Sigma %*% t(Sigma)
  
  Omega_h <- tryCatch(
    omega_h(A, Q, h),
    error = function(e) NULL
  )
  
  Omega_3h2 <- tryCatch(
    omega_h(A, Q, (3 / 2) * h),
    error = function(e) NULL
  )
  
  Omega_RR_h <- Omega_h[2, 2]
  Omega_RR_3h2 <- Omega_3h2[2, 2]
  
  n_terms <- N - 2
  terms <- numeric(n_terms)
  
  for (k in 1:n_terms) {
    z_old <- Y[k, ]
    z_new <- Y[k + 1, ]
 
    z_start <- f_flow(z_old, h / 2, theta, lambda, bx)
    

    m <- mu_h(z_start, A, b_vec, h)
    
    y <- f_flow_inv(z_new, h / 2, theta, lambda, bx)
    
    r_R <- y[2] - m[2]
    
    jac <- logdet_Dv_f_forward(z_new, h / 2, theta, lambda, bx)
    
    terms[k] <- r_R^2 / Omega_RR_h +
      (2 / 3) * log(Omega_RR_3h2) +
      2 * jac
  }
  
  value <- sum(terms)
  
  if (return_terms) return(terms)
  
  value
}
```

#now we test it for what we will call model M1 AND we fit it to the real greenland data
#b=d=alpha=beta=0
```{r}
ss_negloglik_M1_partial_R <- function(
    par,
    x_obs,
    h,
    branch = "+",
    return_terms = FALSE
) {
  
  par_full <- c(
    unname(par["eta"]),
    unname(par["a"]),
    0,
    unname(par["c"]),
    0,
    unname(par["gamma"])
  )
  
  names(par_full) <- c("eta", "a", "b", "c", "d", "gamma")
  
  ss_negloglik_partial_R(
    par = par_full,
    x_obs = x_obs,
    h = h,
    branch = branch,
    return_terms = return_terms
  )
}

Ca2 <- Ca2_final$X
hCa <- 0.02

theta_start_M1 <- c(
  eta   = 50,
  a     = -100,
  c     = 250,
  gamma = sqrt(8800)
)

eps <- 10^-5
lower_bounds_M1 <- c(
  eta   = eps,
  a     = -Inf,
  c     = eps,
  sigma = eps
)

upper_bounds_M1 <- c(
  eta   = Inf,
  a     = -eps,
  c     = Inf,
  sigma = Inf
)

#we test it before optimizing
ss_negloglik_M1_partial_R(
  par = theta_start_M1,
  x_obs = Ca2,
  h = hCa,
  branch = "+"
)


fit_M1_plus <- optim(
  par = theta_start_M1,
  fn = ss_negloglik_M1_partial_R,
  x_obs = Ca2,
  h = hCa,
  branch = "+",
  method = "L-BFGS-B",
  lower = lower_bounds_M1,
  upper = upper_bounds_M1,
  control = list(
    maxit = 1000,
    trace = 1
  )
)

fit_M1_minus <- optim(
  par = theta_start_M1,
  fn = ss_negloglik_M1_partial_R,
  x_obs = Ca2,
  h = hCa,
  branch = "-",
  method = "L-BFGS-B",
  lower = lower_bounds_M1,
  upper = upper_bounds_M1,
  control = list(
    maxit = 1000,
    trace = 1
  )
)

fit_M1_plus$par
fit_M1_plus$value
fit_M1_plus$convergence

fit_M1_minus$par
fit_M1_minus$value
fit_M1_minus$convergence
```



#Now we move on to NON constant noise
#First we define again the required functions (some are the same used in the 
#constant and simulation setting)

#Here we have implemented both methods approached in the report for computing the covariance
#(method 1 was the one I tried to develop and method 2 the one used in the paper)
```{r}
#just a helper function
is_bad <- function(x) {
  any(!is.finite(x))
}

compute_bx <- function(theta, x_data, branch = "+") {
  a <- theta["a"]
  b <- theta["b"]
  
  Ex <- mean(x_data)
  Vx <- var(x_data)
  
  disc <- Vx + (Ex + b / (3 * a))^2
  
  if (branch == "+") {
    -b / (3 * a) + sqrt(disc)
  } else if (branch == "-") {
    -b / (3 * a) - sqrt(disc)
  }
}

make_A_matrix <- function(theta, x_data, branch = "+") {
  eta <- theta["eta"]
  a   <- theta["a"]
  b   <- theta["b"]
  c   <- theta["c"]
  
  Ex  <- mean(x_data)
  Ex2 <- mean(x_data^2)
  
  lambda <- 3 * a * Ex2 + 2 * b * Ex + c
  bx <- compute_bx(theta, x_data, branch = branch)
  
  A <- matrix(
    c(0,       1,
      lambda, -eta),
    nrow = 2,
    byrow = TRUE
  )
  
  b_vec <- c(bx, 0)
  
  list(
    A = A,
    b_vec = b_vec,
    lambda = lambda,
    bx = bx
  )
}

N_fun <- function(state, theta, lambda, bx) {
  x <- state[1]
  
  a <- theta["a"]
  b <- theta["b"]
  c <- theta["c"]
  d <- theta["d"]
  
  n2 <- a * x^3 + b * x^2 + c * x + d - lambda * (x - bx)
  
  c(0, n2)
}

f_flow <- function(state, t, theta, lambda, bx) {
  state + t * N_fun(state, theta, lambda, bx)
}

f_flow_inv <- function(state, t, theta, lambda, bx) {
  state - t * N_fun(state, theta, lambda, bx)
}

logdet_Dv <- function(state, t, theta, lambda, bx) {
  0
}

make_data_XV <- function(x_obs, h) {
  cbind(
    x = x_obs[-length(x_obs)],
    v = diff(x_obs) / h
  )
}

make_par_tilde <- function(diff_par) {
  alpha <- diff_par["alpha"]
  beta  <- diff_par["beta"]
  gamma <- diff_par["gamma"]
  
  alpha_tilde <- matrix(0, 4, 4)
  beta_tilde  <- matrix(0, 4, 2)
  gamma_vec   <- c(0, 0, 0, gamma)
  
  # vectorization:
  # vec([[q11, q12],
  #      [q21, q22]]) = c(q11, q21, q12, q22)
  #
  # Only q22 = alpha*v^2 + beta*v + gamma is nonzero.
  alpha_tilde[4, 4] <- alpha
  beta_tilde[4, 2]  <- beta
  
  list(
    alpha_tilde = alpha_tilde,
    beta_tilde = beta_tilde,
    gamma_vec = gamma_vec
  )
}

Cap_sigma <- function(y, diff_par) {
  v <- y[2]
  
  alpha <- diff_par["alpha"]
  beta  <- diff_par["beta"]
  gamma <- diff_par["gamma"]
  
  q22 <- alpha * v^2 + beta * v + gamma
  
  matrix(
    c(0,   0,
      0, q22),
    nrow = 2,
    byrow = TRUE
  )
}


# Method 1: system u' = B u

build_B <- function(A, b_vec, diff_par) {
  tilde <- make_par_tilde(diff_par)
  
  alpha_tilde <- tilde$alpha_tilde
  beta_tilde  <- tilde$beta_tilde
  gamma_vec   <- tilde$gamma_vec
  
  a0 <- as.numeric(-A %*% b_vec)
  
  K <- kronecker(diag(2), A) + kronecker(A, diag(2))
  
  C_a0m <- matrix(0, 4, 2)
  
  for (j in 1:2) {
    ej <- rep(0, 2)
    ej[j] <- 1
    
    C_a0m[, j] <- as.vector(
      a0 %*% t(ej) + ej %*% t(a0)
    )
  }
  
  B <- matrix(0, 7, 7)
  
  # m' = A m + a0
  B[2:3, 1] <- a0
  B[2:3, 2:3] <- A
  
  # vec(S)' = gamma + (C_a0m + beta_tilde)m + (K + alpha_tilde)vec(S)
  B[4:7, 1] <- gamma_vec
  B[4:7, 2:3] <- C_a0m + beta_tilde
  B[4:7, 4:7] <- K + alpha_tilde
  
  B
}

prepare_omega_B <- function(A, b_vec, diff_par, h) {
  B <- build_B(A, b_vec, diff_par)
  
  list(
    method = "raw",
    eBh = expm(B * h),
    eAh = expm(A * h),
    b_vec = b_vec
  )
}

omega_B_eval <- function(pre, x0) {
  S0 <- x0 %*% t(x0)
  
  u0 <- c(
    1,
    x0,
    as.vector(S0)
  )
  
  uh <- as.numeric(pre$eBh %*% u0)
  
  m_h <- uh[2:3]
  S_h <- matrix(uh[4:7], nrow = 2, ncol = 2)
  
  Omega <- S_h - m_h %*% t(m_h)
  Omega <- 0.5 * (Omega + t(Omega))
  
  Omega
}


# Method 2: I1,...,I5 used in the covariance computation

ap_integral <- function(M, N, P, h) {
  r <- nrow(M)
  c <- nrow(P)
  
  B <- rbind(
    cbind(M, N),
    cbind(matrix(0, c, r), P)
  )
  
  E <- expm(B * h)
  
  E[1:r, (r + 1):(r + c)]
}

compute_I_matrices <- function(A, alpha_tilde, beta_tilde, h) {
  d <- nrow(A)
  d2 <- d * d
  
  K <- kronecker(diag(d), A) + kronecker(A, diag(d))
  M <- K + alpha_tilde
  
  I_kron_A <- kronecker(diag(d), A)
  A_kron_I <- kronecker(A, diag(d))
  
  I1 <- ap_integral(
    M = M,
    N = alpha_tilde,
    P = K,
    h = h
  )
  
  I2 <- ap_integral(
    M = M,
    N = alpha_tilde,
    P = I_kron_A,
    h = h
  )
  
  I3 <- ap_integral(
    M = M,
    N = alpha_tilde,
    P = A_kron_I,
    h = h
  )
  
  I4 <- ap_integral(
    M = M,
    N = beta_tilde,
    P = A,
    h = h
  )
  
  I5 <- ap_integral(
    M = M,
    N = diag(d2),
    P = matrix(0, d2, d2),
    h = h
  )
  
  list(
    I1 = I1,
    I2 = I2,
    I3 = I3,
    I4 = I4,
    I5 = I5
  )
}

prepare_omega_I <- function(A, b_vec, diff_par, h) {
  tilde <- make_par_tilde(diff_par)
  
  I_mats <- compute_I_matrices(
    A = A,
    alpha_tilde = tilde$alpha_tilde,
    beta_tilde = tilde$beta_tilde,
    h = h
  )
  
  Sigma_b <- Cap_sigma(b_vec, diff_par)
  
  list(
    method = "I",
    I = I_mats,
    eAh = expm(A * h),
    b_vec = b_vec,
    Sigma_b_vec = as.vector(Sigma_b)
  )
}

omega_I_eval <- function(pre, x0) {
  I <- pre$I
  b_vec <- pre$b_vec
  
  xb <- as.numeric(x0 - b_vec)
  
  term1 <- I$I1 %*% as.vector(xb %*% t(xb))
  term2 <- I$I2 %*% as.vector(xb %*% t(b_vec))
  term3 <- I$I3 %*% as.vector(b_vec %*% t(xb))
  term4 <- I$I4 %*% xb
  term5 <- I$I5 %*% pre$Sigma_b_vec
  
  vecOmega <- term1 + term2 + term3 + term4 + term5
  
  Omega <- matrix(vecOmega, nrow = 2, ncol = 2)
  Omega <- 0.5 * (Omega + t(Omega))
  
  Omega
}

# Here one can chose the covariance computation method 
prepare_omega <- function(
    A,
    b_vec,
    diff_par,
    h,
    method = c("raw", "I")
) {
  method <- match.arg(method)
  
  if (method == "raw") {
    prepare_omega_B(A, b_vec, diff_par, h)
  } else {
    prepare_omega_I(A, b_vec, diff_par, h)
  }
}

omega_eval <- function(pre, x0) {
  if (pre$method == "raw") {
    omega_B_eval(pre, x0)
  } else if (pre$method == "I") {
    omega_I_eval(pre, x0)
  }
}
```


#We are now able to compute the objective function
```{r}
ss_negloglik_nonC <- function(
    par,
    x_obs,
    h,
    branch = "+",
    omega_method = c("raw", "I"),
    invalid = Inf,
    return_terms = FALSE
) {
  omega_method <- match.arg(omega_method)
  
  theta <- par[c("eta", "a", "b", "c", "d")]
  diff_par <- par[c("alpha", "beta", "gamma")]
  
  N <- length(x_obs)
  
  Y <- cbind(
    x = x_obs[1:(N - 1)],
    v = (x_obs[2:N] - x_obs[1:(N - 1)]) / h
  )
  
  if (is_bad(Y)) return(invalid)
  
  x_data <- Y[, "x"]
  
  lin <- make_A_matrix(theta, x_data, branch = branch)
  if (is.null(lin)) return(invalid)
  
  A      <- lin$A
  b_vec  <- lin$b_vec
  lambda <- lin$lambda
  bx     <- lin$bx
  
  if (is_bad(A) || is_bad(b_vec)) return(invalid)
  
  pre_h <- tryCatch(
    prepare_omega(
      A = A,
      b_vec = b_vec,
      diff_par = diff_par,
      h = h,
      method = omega_method
    ),
    error = function(e) NULL,
    warning = function(w) NULL
  )
  
  pre_3h2 <- tryCatch(
    prepare_omega(
      A = A,
      b_vec = b_vec,
      diff_par = diff_par,
      h = (3 / 2) * h,
      method = omega_method
    ),
    error = function(e) NULL,
    warning = function(w) NULL
  )
  
  if (is.null(pre_h) || is.null(pre_3h2)) return(invalid)
  
  n_terms <- N - 2
  terms <- numeric(n_terms)
  
  for (k in seq_len(n_terms)) {
    z_old <- Y[k, ]
    z_new <- Y[k + 1, ]
    
    z_start <- f_flow(z_old, h / 2, theta, lambda, bx)
    if (is_bad(z_start)) return(invalid)
    
    m <- as.numeric(pre_h$eAh %*% (z_start - b_vec) + b_vec)
    if (is_bad(m)) return(invalid)
    
    y <- f_flow_inv(z_new, h / 2, theta, lambda, bx)
    if (is_bad(y)) return(invalid)
    
    r_R <- y[2] - m[2]
    if (!is.finite(r_R)) return(invalid)
    
    Omega_h <- tryCatch(
      omega_eval(pre_h, z_start),
      error = function(e) NULL,
      warning = function(w) NULL
    )
    
    Omega_3h2 <- tryCatch(
      omega_eval(pre_3h2, z_start),
      error = function(e) NULL,
      warning = function(w) NULL
    )
    
    if (is.null(Omega_h) || is.null(Omega_3h2)) return(invalid)
    if (is_bad(Omega_h) || is_bad(Omega_3h2)) return(invalid)
    
    Omega_RR_h <- Omega_h[2, 2]
    Omega_RR_3h2 <- Omega_3h2[2, 2]
    
    if (!is.finite(Omega_RR_h) || Omega_RR_h <= 0) return(invalid)
    if (!is.finite(Omega_RR_3h2) || Omega_RR_3h2 <= 0) return(invalid)
    
    jac <- logdet_Dv(z_new, h / 2, theta, lambda, bx)
    if (!is.finite(jac)) return(invalid)
    
    terms[k] <- r_R^2 / Omega_RR_h +
      (2 / 3) * log(Omega_RR_3h2) +
      2 * jac
    
    if (!is.finite(terms[k])) return(invalid)
  }
  
  value <- sum(terms)
  
  if (!is.finite(value)) return(invalid)
  
  if (return_terms) return(terms)
  
  value
}
```

#We turn to model 2
```{r}
# M2 : b = 0, d = 0

ss_negloglik_M2_partial_R <- function(
    par,
    x_obs,
    h,
    branch = "+",
    omega_method = c("raw", "I"),
    invalid = Inf,
    return_terms = FALSE
) {
  omega_method <- match.arg(omega_method)
  
  par_full <- c(
    eta   = unname(par["eta"]),
    a     = unname(par["a"]),
    b     = 0,
    c     = unname(par["c"]),
    d     = 0,
    alpha = unname(par["alpha"]),
    beta  = unname(par["beta"]),
    gamma = unname(par["gamma"])
  )
  
  ss_negloglik_nonC(
    par = par_full,
    x_obs = x_obs,
    h = h,
    branch = branch,
    omega_method = omega_method,
    invalid = invalid,
    return_terms = return_terms
  )
}

objective_M2_for_optim <- function(
    par,
    x_obs,
    h,
    branch = "+",
    omega_method = c("raw", "I")
) {
  omega_method <- match.arg(omega_method)
  
  invalid_finite <- 1e100
  
  val <- tryCatch(
    ss_negloglik_M2_partial_R(
      par = par,
      x_obs = x_obs,
      h = h,
      branch = branch,
      omega_method = omega_method,
      invalid = invalid_finite
    ),
    error = function(e) invalid_finite,
    warning = function(w) invalid_finite
  )
  
  if (!is.finite(val)) invalid_finite else val
}


Ca2 <- Ca2_final$X
hCa <- 0.02

theta_start_M2 <- c(
  eta   = 67.9,
  a     = -96.6,
  c     = 65.4,
  alpha = 64.3,
  beta  = 233.2,
  gamma = 4387.8
)

lower_bounds_M2 <- c(
  eta   = 1e-8,
  a     = -Inf,
  c     = 1e-8,
  alpha = 1e-8,
  beta  = -Inf,
  gamma = 1e-8
)

upper_bounds_M2 <- c(
  eta   = Inf,
  a     = -1e-8,
  c     = Inf,
  alpha = Inf,
  beta  = Inf,
  gamma = Inf
)

# Testing 
ss_negloglik_M2_partial_R(
  par = theta_start_M2,
  x_obs = Ca2,
  h = hCa,
  branch = "+",
  omega_method = "raw"
)

# Optimization

fit_M2_raw_plus <- optim(
  par = theta_start_M2,
  fn = objective_M2_for_optim,
  x_obs = Ca2,
  h = hCa,
  branch = "+",
  omega_method = "raw",
  method = "L-BFGS-B",
  lower = lower_bounds_M2,
  upper = upper_bounds_M2,
  control = list(maxit = 1000, trace = 1)
)

fit_M2_raw_minus <- optim(
  par = theta_start_M2,
  fn = objective_M2_for_optim,
  x_obs = Ca2,
  h = hCa,
  branch = "-",
  omega_method = "raw",
  method = "L-BFGS-B",
  lower = lower_bounds_M2,
  upper = upper_bounds_M2,
  control = list(maxit = 1000, trace = 1)
)

fit_M2_I_plus <- optim(
  par = theta_start_M2,
  fn = objective_M2_for_optim,
  x_obs = Ca2,
  h = hCa,
  branch = "+",
  omega_method = "I",
  method = "L-BFGS-B",
  lower = lower_bounds_M2,
  upper = upper_bounds_M2,
  control = list(maxit = 1000, trace = 1)
)

fit_M2_I_minus <- optim(
  par = theta_start_M2,
  fn = objective_M2_for_optim,
  x_obs = Ca2,
  h = hCa,
  branch = "-",
  omega_method = "I",
  method = "L-BFGS-B",
  lower = lower_bounds_M2,
  upper = upper_bounds_M2,
  control = list(maxit = 1000, trace = 1)
)
```


#We can now compare the covariances in both methods, and the computation time
#for the covariance, objective function and optimization 

#This is done exactly the same way for model 3
```{r}
check_omega_methods <- function(par, x_obs, h, branch = "+") {
  par_full <- c(
    eta   = unname(par["eta"]),
    a     = unname(par["a"]),
    b     = 0,
    c     = unname(par["c"]),
    d     = 0,
    alpha = unname(par["alpha"]),
    beta  = unname(par["beta"]),
    gamma = unname(par["gamma"])
  )
  
  theta <- par_full[c("eta", "a", "b", "c", "d")]
  diff_par <- par_full[c("alpha", "beta", "gamma")]
  
  Y <- make_data_XV(x_obs, h)
  lin <- make_A_matrix(theta, Y[, "x"], branch = branch)
  
  A <- lin$A
  b_vec <- lin$b_vec
  lambda <- lin$lambda
  bx <- lin$bx
  
  z_old <- Y[1, ]
  z_start <- f_flow(z_old, h / 2, theta, lambda, bx)
  
  pre_raw <- prepare_omega(A, b_vec, diff_par, h, method = "raw")
  pre_I <- prepare_omega(A, b_vec, diff_par, h, method = "I")
  
  Omega_raw <- omega_eval(pre_raw, z_start)
  Omega_I <- omega_eval(pre_I, z_start)
  
  list(
    Omega_raw = Omega_raw,
    Omega_I = Omega_I,
    difference = Omega_raw - Omega_I,
    max_abs_difference = max(abs(Omega_raw - Omega_I))
  )
}

check_omega_methods(
  par = theta_start_M2,
  x_obs = Ca2,
  h = hCa,
  branch = "+"
)


#NOW COMPARING THE COMPUTATION TIME
time_expr <- function(expr) {
  t0 <- proc.time()
  result <- eval.parent(substitute(expr))
  elapsed <- proc.time() - t0
  
  list(
    result = result,
    elapsed = elapsed["elapsed"]
  )
}


benchmark_omega_methods <- function(
    par,
    x_obs,
    h,
    branch = "+",
    n_states = 500
) {
  
  par_full <- c(
    eta   = unname(par["eta"]),
    a     = unname(par["a"]),
    b     = 0,
    c     = unname(par["c"]),
    d     = 0,
    alpha = unname(par["alpha"]),
    beta  = unname(par["beta"]),
    gamma = unname(par["gamma"])
  )
  
  theta <- par_full[c("eta", "a", "b", "c", "d")]
  diff_par <- par_full[c("alpha", "beta", "gamma")]
  
  Y <- make_data_XV(x_obs, h)
  
  lin <- make_A_matrix(theta, Y[, "x"], branch = branch)
  
  A <- lin$A
  b_vec <- lin$b_vec
  lambda <- lin$lambda
  bx <- lin$bx
  
  n_available <- nrow(Y) - 1
  idx <- round(seq(1, n_available, length.out = min(n_states, n_available)))
  
  Z_start <- t(sapply(idx, function(k) {
    f_flow(Y[k, ], h / 2, theta, lambda, bx)
  }))
  
  # Precompute once for each method
  pre_raw <- prepare_omega(
    A = A,
    b_vec = b_vec,
    diff_par = diff_par,
    h = h,
    method = "raw"
  )
  
  pre_I <- prepare_omega(
    A = A,
    b_vec = b_vec,
    diff_par = diff_par,
    h = h,
    method = "I"
  )
  
  # Time method 1 
  raw_time <- time_expr({
    out_raw <- vector("list", nrow(Z_start))
    for (i in seq_len(nrow(Z_start))) {
      out_raw[[i]] <- omega_eval(pre_raw, Z_start[i, ])
    }
    out_raw
  })
  
  # Time method 2
  I_time <- time_expr({
    out_I <- vector("list", nrow(Z_start))
    for (i in seq_len(nrow(Z_start))) {
      out_I[[i]] <- omega_eval(pre_I, Z_start[i, ])
    }
    out_I
  })
  
  # Compare one omega to make sure both are not wildly different
  Omega_raw_1 <- omega_eval(pre_raw, Z_start[1, ])
  Omega_I_1 <- omega_eval(pre_I, Z_start[1, ])
  
  list(
    n_states = nrow(Z_start),
    raw_seconds = raw_time$elapsed,
    I_seconds = I_time$elapsed,
    speed_ratio_I_over_raw = I_time$elapsed / raw_time$elapsed,
    Omega_raw_first = Omega_raw_1,
    Omega_I_first = Omega_I_1,
    max_abs_difference_first = max(abs(Omega_raw_1 - Omega_I_1))
  )
}

bench_omega_plus <- benchmark_omega_methods(
  par = theta_start_M2,
  x_obs = Ca2,
  h = hCa,
  branch = "+",
  n_states = 1000
)

bench_omega_minus <- benchmark_omega_methods(
  par = theta_start_M2,
  x_obs = Ca2,
  h = hCa,
  branch = "-",
  n_states = 1000
)

bench_omega_plus
bench_omega_minus

#Interpretation:
#speed_ratio_I_over_raw > 1
#means "method 2" is slower than "method 1"

benchmark_likelihood_methods <- function(
    par,
    x_obs,
    h,
    branch = "+",
    n_rep = 5
) {
  
  raw_time <- time_expr({
    raw_values <- numeric(n_rep)
    for (i in seq_len(n_rep)) {
      raw_values[i] <- ss_negloglik_M2_partial_R(
        par = par,
        x_obs = x_obs,
        h = h,
        branch = branch,
        omega_method = "raw"
      )
    }
    raw_values
  })
  
  I_time <- time_expr({
    I_values <- numeric(n_rep)
    for (i in seq_len(n_rep)) {
      I_values[i] <- ss_negloglik_M2_partial_R(
        par = par,
        x_obs = x_obs,
        h = h,
        branch = branch,
        omega_method = "I"
      )
    }
    I_values
  })
  
  list(
    branch = branch,
    n_rep = n_rep,
    raw_seconds_total = raw_time$elapsed,
    I_seconds_total = I_time$elapsed,
    raw_seconds_per_eval = raw_time$elapsed / n_rep,
    I_seconds_per_eval = I_time$elapsed / n_rep,
    speed_ratio_I_over_raw = I_time$elapsed / raw_time$elapsed,
    raw_values = raw_time$result,
    I_values = I_time$result
  )
}

bench_lik_plus <- benchmark_likelihood_methods(
  par = theta_start_M2,
  x_obs = Ca2,
  h = hCa,
  branch = "+",
  n_rep = 5
)

bench_lik_minus <- benchmark_likelihood_methods(
  par = theta_start_M2,
  x_obs = Ca2,
  h = hCa,
  branch = "-",
  n_rep = 5
)

bench_lik_plus
bench_lik_minus


#One can also compare the optimization times
benchmark_optim_methods <- function(
    par_start,
    x_obs,
    h,
    branch = "+",
    lower,
    upper,
    maxit = 300
) {
  
  raw_fit_time <- time_expr({
    optim(
      par = par_start,
      fn = objective_M2_for_optim,
      x_obs = x_obs,
      h = h,
      branch = branch,
      omega_method = "raw",
      method = "L-BFGS-B",
      lower = lower,
      upper = upper,
      control = list(
        maxit = maxit,
        trace = 0
      )
    )
  })
  
  I_fit_time <- time_expr({
    optim(
      par = par_start,
      fn = objective_M2_for_optim,
      x_obs = x_obs,
      h = h,
      branch = branch,
      omega_method = "I",
      method = "L-BFGS-B",
      lower = lower,
      upper = upper,
      control = list(
        maxit = maxit,
        trace = 0
      )
    )
  })
  
  raw_fit <- raw_fit_time$result
  I_fit <- I_fit_time$result
  
  list(
    branch = branch,
    maxit = maxit,
    
    raw_seconds = raw_fit_time$elapsed,
    raw_par = raw_fit$par,
    raw_value = raw_fit$value,
    raw_convergence = raw_fit$convergence,
    raw_counts = raw_fit$counts,
    
    I_seconds = I_fit_time$elapsed,
    I_par = I_fit$par,
    I_value = I_fit$value,
    I_convergence = I_fit$convergence,
    I_counts = I_fit$counts,
    
    speed_ratio_I_over_raw = I_fit_time$elapsed / raw_fit_time$elapsed
  )
}

bench_optim_plus <- benchmark_optim_methods(
  par_start = theta_start_M2,
  x_obs = Ca2,
  h = hCa,
  branch = "+",
  lower = lower_bounds_M2,
  upper = upper_bounds_M2,
  maxit = 200
)

bench_optim_minus <- benchmark_optim_methods(
  par_start = theta_start_M2,
  x_obs = Ca2,
  h = hCa,
  branch = "-",
  lower = lower_bounds_M2,
  upper = upper_bounds_M2,
  maxit = 200
)

bench_optim_plus
bench_optim_minus
```


#We now move to model 3
```{r}
# M3 : estimates all parameters

ss_negloglik_M3_partial_R <- function(
    par,
    x_obs,
    h,
    branch = "+",
    omega_method = c("raw", "I"),
    invalid = Inf,
    return_terms = FALSE
) {
  omega_method <- match.arg(omega_method)
  
  ss_negloglik_nonC(
    par = par,
    x_obs = x_obs,
    h = h,
    branch = branch,
    omega_method = omega_method,
    invalid = invalid,
    return_terms = return_terms
  )
}

objective_M3_for_optim <- function(
    par,
    x_obs,
    h,
    branch = "+",
    omega_method = c("raw", "I")
) {
  omega_method <- match.arg(omega_method)
  
  invalid_finite <- 1e100
  
  val <- tryCatch(
    ss_negloglik_M3_partial_R(
      par = par,
      x_obs = x_obs,
      h = h,
      branch = branch,
      omega_method = omega_method,
      invalid = invalid_finite
    ),
    error = function(e) invalid_finite,
    warning = function(w) invalid_finite
  )
  
  if (!is.finite(val)) invalid_finite else val
}

theta_start_M3 <- c(
  eta   = 67.9,
  a     = -97.8,
  b     = 8.4,
  c     = 66.2,
  d     = -9.1,
  alpha = 64.4,
  beta  = 228.0,
  gamma = 4371.0
)

eps <- 1e-8

lower_bounds_M3 <- c(
  eta   = eps,
  a     = -Inf,
  b     = -Inf,
  c     = -Inf,
  d     = -Inf,
  alpha = eps,
  beta  = -Inf,
  gamma = eps
)

upper_bounds_M3 <- c(
  eta   = Inf,
  a     = -eps,
  b     = Inf,
  c     = Inf,
  d     = Inf,
  alpha = Inf,
  beta  = Inf,
  gamma = Inf
)

#Optimization with method 1
fit_M3_raw_plus <- optim(
  par = theta_start_M3,
  fn = objective_M3_for_optim,
  x_obs = Ca2,
  h = hCa,
  branch = "+",
  omega_method = "raw",
  method = "L-BFGS-B",
  lower = lower_bounds_M3,
  upper = upper_bounds_M3,
  control = list(
    maxit = 1000,
    trace = 1
  )
)

fit_M3_raw_minus <- optim(
  par = theta_start_M3,
  fn = objective_M3_for_optim,
  x_obs = Ca2,
  h = hCa,
  branch = "-",
  omega_method = "raw",
  method = "L-BFGS-B",
  lower = lower_bounds_M3,
  upper = upper_bounds_M3,
  control = list(
    maxit = 1000,
    trace = 1
  )
)

# Optimization with method 2 

fit_M3_I_plus <- optim(
  par = theta_start_M3,
  fn = objective_M3_for_optim,
  x_obs = Ca2,
  h = hCa,
  branch = "+",
  omega_method = "I",
  method = "L-BFGS-B",
  lower = lower_bounds_M3,
  upper = upper_bounds_M3,
  control = list(
    maxit = 1000,
    trace = 1
  )
)

fit_M3_I_minus <- optim(
  par = theta_start_M3,
  fn = objective_M3_for_optim,
  x_obs = Ca2,
  h = hCa,
  branch = "-",
  omega_method = "I",
  method = "L-BFGS-B",
  lower = lower_bounds_M3,
  upper = upper_bounds_M3,
  control = list(
    maxit = 1000,
    trace = 1
  )
)
```


#Estimates comparison between methods and models used
```{r}
make_fit_row <- function(fit, model_name, method, branch, has_b_d = TRUE) {
  
  p <- fit$par
  
  data.frame(
    model = model_name,
    method = method,
    branch = branch,
    eta = unname(p["eta"]),
    a = unname(p["a"]),
    b = if ("b" %in% names(p)) unname(p["b"]) else 0,
    c = unname(p["c"]),
    d = if ("d" %in% names(p)) unname(p["d"]) else 0,
    alpha = unname(p["alpha"]),
    beta = unname(p["beta"]),
    gamma = unname(p["gamma"]),
    value = fit$value,
    convergence = fit$convergence,
    row.names = NULL
  )
}

comparison_fits <- rbind(
  make_fit_row(fit_M2_raw_plus,  "M2", "raw", "+"),
  make_fit_row(fit_M2_raw_minus, "M2", "raw", "-"),
  make_fit_row(fit_M2_I_plus,    "M2", "I",   "+"),
  make_fit_row(fit_M2_I_minus,   "M2", "I",   "-"),
  make_fit_row(fit_M3_raw_plus,  "M3", "raw", "+"),
  make_fit_row(fit_M3_raw_minus, "M3", "raw", "-"),
  make_fit_row(fit_M3_I_plus,    "M3", "I",   "+"),
  make_fit_row(fit_M3_I_minus,   "M3", "I",   "-")
)

comparison_fits
```


#BOOTSTRAP CONFIDENCE INTERVALS FOR MODEL 3, PARAMETERS B AND D, bx root with +
```{r}
expand_M3_par <- function(par) {
  c(
    eta   = unname(par["eta"]),
    a     = unname(par["a"]),
    b     = unname(par["b"]),
    c     = unname(par["c"]),
    d     = unname(par["d"]),
    alpha = unname(par["alpha"]),
    beta  = unname(par["beta"]),
    gamma = unname(par["gamma"])
  )
}

simulate_bootstrap_X <- function(
    par,
    N_obs,
    h_obs,
    h_sim = 0.005,
    burn_obs = 200,
    start = c(0, 0)
) {
  par <- expand_M3_par(par)
  
  eta   <- par["eta"]
  a     <- par["a"]
  b     <- par["b"]
  c     <- par["c"]
  d     <- par["d"]
  alpha <- par["alpha"]
  beta  <- par["beta"]
  gamma <- par["gamma"]
  
  ratio <- h_obs / h_sim
  skip <- round(ratio)
  
  N_total_obs <- N_obs + burn_obs
  N_fine <- (N_total_obs - 1) * skip + 1
  
  x <- start[1]
  v <- start[2]
  
  X_obs <- numeric(N_total_obs)
  obs_counter <- 1
  X_obs[obs_counter] <- x
  
  for (i in 2:N_fine) {
    drift_v <- -eta * v +
      a * x^3 +
      b * x^2 +
      c * x +
      d
    
    diff2 <- alpha * v^2 + beta * v + gamma
    diff2 <- max(diff2, 1e-12)
    
    eps <- rnorm(1)
    
    x_new <- x + h_sim * v
    v_new <- v + h_sim * drift_v + sqrt(diff2) * sqrt(h_sim) * eps
    
    x <- x_new
    v <- v_new
    
    if ((i - 1) %% skip == 0) {
      obs_counter <- obs_counter + 1
      X_obs[obs_counter] <- x
    }
  }
  
  X_obs[(burn_obs + 1):(burn_obs + N_obs)]
}

fit_M3_I_plus_safe <- function(
    x_boot,
    h,
    start_par,
    lower,
    upper,
    maxit = 300,
    factr = 1e9
) {
  invalid_finite <- 1e100
  
  # Check objective at starting point first
  start_val <- objective_M3_for_optim(
    par = start_par,
    x_obs = x_boot,
    h = h,
    branch = "+",
    omega_method = "I"
  )
  
  if (!is.finite(start_val) || start_val >= 0.99 * invalid_finite) {
    return(NULL)
  }
  
  fit <- tryCatch(
    optim(
      par = start_par,
      fn = objective_M3_for_optim,
      x_obs = x_boot,
      h = h,
      branch = "+",
      omega_method = "I",
      method = "L-BFGS-B",
      lower = lower,
      upper = upper,
      control = list(
        maxit = maxit,
        trace = 0,
        factr = factr
      )
    ),
    error = function(e) NULL
  )
  
  if (is.null(fit)) return(NULL)
  
  if (!is.finite(fit$value) || fit$value >= 0.99 * invalid_finite) {
    return(NULL)
  }
  
  fit
}

bootstrap_row_M3_safe <- function(fit, idx, elapsed, start_val = NA_real_) {
  if (is.null(fit)) {
    return(data.frame(
      traj = idx,
      nll = NA_real_,
      start_nll = start_val,
      time_sec = elapsed,
      convergence = NA_integer_,
      eta = NA_real_,
      a = NA_real_,
      b = NA_real_,
      c = NA_real_,
      d = NA_real_,
      alpha = NA_real_,
      beta = NA_real_,
      gamma = NA_real_
    ))
  }
  
  p <- expand_M3_par(fit$par)
  
  data.frame(
    traj = idx,
    nll = fit$value,
    start_nll = start_val,
    time_sec = elapsed,
    convergence = fit$convergence,
    eta = p["eta"],
    a = p["a"],
    b = p["b"],
    c = p["c"],
    d = p["d"],
    alpha = p["alpha"],
    beta = p["beta"],
    gamma = p["gamma"],
    row.names = NULL
  )
}


bootstrap_M3_I_plus <- function(
    fit_M3_I_plus,
    x_obs,
    h,
    B = 50,
    h_sim = 0.005,
    burn_obs = 200,
    lower = lower_bounds_M3,
    upper = upper_bounds_M3,
    seed = 123,
    maxit = 300,
    results_file = "bootstrap_M3_I_plus_test.csv"
) {
  par_hat <- expand_M3_par(fit_M3_I_plus$par)
  N_obs <- length(x_obs)
  
  if (file.exists(results_file)) {
    old <- read.csv(results_file)
    
    if (nrow(old) > 0) {
      start_idx <- max(old$traj, na.rm = TRUE) + 1
    } else {
      start_idx <- 1
    }
    
    cat("Resuming from bootstrap replicate", start_idx, "\n")
  } else {
    start_idx <- 1
    cat("Starting new bootstrap\n")
  }
  
  if (start_idx > B) {
    cat("Bootstrap already completed.\n")
    return(read.csv(results_file))
  }
  
  for (bb in start_idx:B) {
    cat("Bootstrap", bb, "of", B, "\n")
    
    set.seed(seed + bb)
    
    t0 <- proc.time()[["elapsed"]]
    
    x_boot <- simulate_bootstrap_X(
      par = par_hat,
      N_obs = N_obs,
      h_obs = h,
      h_sim = h_sim,
      burn_obs = burn_obs,
      start = c(x_obs[1], 0)
    )
    
    start_val <- objective_M3_for_optim(
      par = par_hat,
      x_obs = x_boot,
      h = h,
      branch = "+",
      omega_method = "I"
    )
    
    cat("Starting NLL:", start_val, "\n")
    
    boot_fit <- fit_M3_I_plus_safe(
      x_boot = x_boot,
      h = h,
      start_par = par_hat,
      lower = lower,
      upper = upper,
      maxit = maxit
    )
    
    elapsed <- proc.time()[["elapsed"]] - t0
    
    row <- bootstrap_row_M3_safe(
      fit = boot_fit,
      idx = bb,
      elapsed = elapsed,
      start_val = start_val
    )
    
    write.table(
      row,
      file = results_file,
      sep = ",",
      row.names = FALSE,
      col.names = !file.exists(results_file),
      append = file.exists(results_file)
    )
    
    gc()
  }
  
  read.csv(results_file)
}

file.remove("bootstrap_M3_I_plus_test.csv")

boot_M3_I_plus <- bootstrap_M3_I_plus(
  fit_M3_I_plus = fit_M3_I_plus,
  x_obs = Ca2,
  h = hCa,
  B = 100,
  h_sim = 0.001,
  burn_obs = 200,
  lower = lower_bounds_M3,
  upper = upper_bounds_M3,
  seed = 123,
  maxit = 300,
  results_file = "bootstrap_M3_I_plus.csv"
)

boot_M3_I_plus <- read.csv("bootstrap_M3_I_plus.csv")

successful_boot <- boot_M3_I_plus[
  is.finite(boot_M3_I_plus$b) &
    is.finite(boot_M3_I_plus$d) &
    boot_M3_I_plus$convergence == 0,
]

nrow(successful_boot)

ci_b <- quantile(successful_boot$b, probs = c(0.025, 0.975), na.rm = TRUE)
ci_d <- quantile(successful_boot$d, probs = c(0.025, 0.975), na.rm = TRUE)

bootstrap_ci_table <- data.frame(
  parameter = c("b", "d"),
  estimate = c(
    fit_M3_I_plus$par["b"],
    fit_M3_I_plus$par["d"]
  ),
  lower_95 = c(ci_b[1], ci_d[1]),
  upper_95 = c(ci_b[2], ci_d[2]),
  contains_zero = c(
    ci_b[1] <= 0 && ci_b[2] >= 0,
    ci_d[1] <= 0 && ci_d[2] >= 0
  ),
  row.names = NULL
)

bootstrap_ci_table
```


#we can also do a parametric bootstrap LR test between m2 and m3 to check which one to chose
```{r}
expand_M2_to_M3_par <- function(par_M2) {
  c(
    eta   = unname(par_M2["eta"]),
    a     = unname(par_M2["a"]),
    b     = 0,
    c     = unname(par_M2["c"]),
    d     = 0,
    alpha = unname(par_M2["alpha"]),
    beta  = unname(par_M2["beta"]),
    gamma = unname(par_M2["gamma"])
  )
}

expand_M3_par <- function(par) {
  c(
    eta   = unname(par["eta"]),
    a     = unname(par["a"]),
    b     = unname(par["b"]),
    c     = unname(par["c"]),
    d     = unname(par["d"]),
    alpha = unname(par["alpha"]),
    beta  = unname(par["beta"]),
    gamma = unname(par["gamma"])
  )
}

simulate_bootstrap_X <- function(
    par,
    N_obs,
    h_obs,
    h_sim = 0.005,
    burn_obs = 200,
    start = c(0, 0)
) {
  par <- expand_M3_par(par)
  
  eta   <- par["eta"]
  a     <- par["a"]
  b     <- par["b"]
  c     <- par["c"]
  d     <- par["d"]
  alpha <- par["alpha"]
  beta  <- par["beta"]
  gamma <- par["gamma"]
  
  ratio <- h_obs / h_sim
  skip <- round(ratio)
  
  
  N_total_obs <- N_obs + burn_obs
  N_fine <- (N_total_obs - 1) * skip + 1
  
  x <- start[1]
  v <- start[2]
  
  X_obs <- numeric(N_total_obs)
  obs_counter <- 1
  X_obs[obs_counter] <- x
  
  for (i in 2:N_fine) {
    drift_v <- -eta * v +
      a * x^3 +
      b * x^2 +
      c * x +
      d
    
    diff2 <- alpha * v^2 + beta * v + gamma
    diff2 <- max(diff2, 1e-12)
    
    eps <- rnorm(1)
    
    x_new <- x + h_sim * v
    v_new <- v + h_sim * drift_v + sqrt(diff2) * sqrt(h_sim) * eps
    
    x <- x_new
    v <- v_new
    
    if ((i - 1) %% skip == 0) {
      obs_counter <- obs_counter + 1
      X_obs[obs_counter] <- x
    }
  }
  
  X_obs[(burn_obs + 1):(burn_obs + N_obs)]
}


fit_M2_I_plus_one_boot <- function(
    x_boot,
    h,
    start_par,
    lower,
    upper,
    maxit = 300,
    factr = 1e9
) {
  invalid_finite <- 1e100
  
  start_val <- objective_M2_for_optim(
    par = start_par,
    x_obs = x_boot,
    h = h,
    branch = "+",
    omega_method = "I"
  )
  
  if (!is.finite(start_val) || start_val >= 0.99 * invalid_finite) {
    return(NULL)
  }
  
  fit <- tryCatch(
    optim(
      par = start_par,
      fn = objective_M2_for_optim,
      x_obs = x_boot,
      h = h,
      branch = "+",
      omega_method = "I",
      method = "L-BFGS-B",
      lower = lower,
      upper = upper,
      control = list(
        maxit = maxit,
        trace = 0,
        factr = factr
      )
    ),
    error = function(e) NULL
  )
  
  if (is.null(fit)) return(NULL)
  if (!is.finite(fit$value) || fit$value >= 0.99 * invalid_finite) return(NULL)
  
  fit
}

fit_M3_I_plus_one_boot <- function(
    x_boot,
    h,
    start_par,
    lower,
    upper,
    maxit = 300,
    factr = 1e9
) {
  invalid_finite <- 1e100
  
  start_val <- objective_M3_for_optim(
    par = start_par,
    x_obs = x_boot,
    h = h,
    branch = "+",
    omega_method = "I"
  )
  
  if (!is.finite(start_val) || start_val >= 0.99 * invalid_finite) {
    return(NULL)
  }
  
  fit <- tryCatch(
    optim(
      par = start_par,
      fn = objective_M3_for_optim,
      x_obs = x_boot,
      h = h,
      branch = "+",
      omega_method = "I",
      method = "L-BFGS-B",
      lower = lower,
      upper = upper,
      control = list(
        maxit = maxit,
        trace = 0,
        factr = factr
      )
    ),
    error = function(e) NULL
  )
  
  if (is.null(fit)) return(NULL)
  if (!is.finite(fit$value) || fit$value >= 0.99 * invalid_finite) return(NULL)
  
  fit
}


bootstrap_LR_row_I_plus <- function(
    idx,
    fit_M2,
    fit_M3,
    elapsed
) {
  if (is.null(fit_M2) || is.null(fit_M3)) {
    return(data.frame(
      traj = idx,
      nll_M2 = NA_real_,
      nll_M3 = NA_real_,
      LR = NA_real_,
      time_sec = elapsed,
      conv_M2 = NA_integer_,
      conv_M3 = NA_integer_
    ))
  }
  
  data.frame(
    traj = idx,
    nll_M2 = fit_M2$value,
    nll_M3 = fit_M3$value,
    LR = 2 * (fit_M2$value - fit_M3$value),
    time_sec = elapsed,
    conv_M2 = fit_M2$convergence,
    conv_M3 = fit_M3$convergence,
    row.names = NULL
  )
}

bootstrap_LR_M2Iplus_vs_M3Iplus <- function(
    fit_M2_I_plus,
    fit_M3_I_plus,
    x_obs,
    h,
    B = 100,
    h_sim = 0.005,
    burn_obs = 200,
    seed = 123,
    maxit_M2 = 300,
    maxit_M3 = 300,
    results_file = "bootstrap_LR_M2Iplus_M3Iplus.csv"
) {
  # Simulate under null model M2
  par_null_full <- expand_M2_to_M3_par(fit_M2_I_plus$par)
  
  # Starting values for optimization
  start_M2 <- fit_M2_I_plus$par
  start_M3 <- fit_M3_I_plus$par
  
  N_obs <- length(x_obs)
  
  if (file.exists(results_file)) {
    old <- read.csv(results_file)
    
    if (nrow(old) > 0) {
      start_idx <- max(old$traj, na.rm = TRUE) + 1
    } else {
      start_idx <- 1
    }
    
    cat("Resuming from bootstrap replicate", start_idx, "\n")
  } else {
    start_idx <- 1
    cat("Starting new LR bootstrap\n")
  }
  
  if (start_idx > B) {
    cat("Bootstrap already completed.\n")
    return(read.csv(results_file))
  }
  
  for (bb in start_idx:B) {
    cat("Bootstrap", bb, "of", B, "\n")
    
    set.seed(seed + bb)
    
    t0 <- proc.time()[["elapsed"]]
    
    # Simulate X-only bootstrap data from fitted M2
    x_boot <- simulate_bootstrap_X(
      par = par_null_full,
      N_obs = N_obs,
      h_obs = h,
      h_sim = h_sim,
      burn_obs = burn_obs,
      start = c(x_obs[1], 0)
    )
    
    fit_M2_b <- fit_M2_I_plus_one_boot(
      x_boot = x_boot,
      h = h,
      start_par = start_M2,
      lower = lower_bounds_M2,
      upper = upper_bounds_M2,
      maxit = maxit_M2
    )
    
    fit_M3_b <- fit_M3_I_plus_one_boot(
      x_boot = x_boot,
      h = h,
      start_par = start_M3,
      lower = lower_bounds_M3,
      upper = upper_bounds_M3,
      maxit = maxit_M3
    )
    
    elapsed <- proc.time()[["elapsed"]] - t0
    
    row <- bootstrap_LR_row_I_plus(
      idx = bb,
      fit_M2 = fit_M2_b,
      fit_M3 = fit_M3_b,
      elapsed = elapsed
    )
    
    write.table(
      row,
      file = results_file,
      sep = ",",
      row.names = FALSE,
      col.names = !file.exists(results_file),
      append = file.exists(results_file)
    )
    
    gc()
  }
  
  read.csv(results_file)
}

file.remove("bootstrap_LR_M2Iplus_M3Iplus_test.csv")

boot_LR_Iplus_test <- bootstrap_LR_M2Iplus_vs_M3Iplus(
  fit_M2_I_plus = fit_M2_I_plus,
  fit_M3_I_plus = fit_M3_I_plus,
  x_obs = Ca2,
  h = hCa,
  B = 5,
  h_sim = 0.001,
  burn_obs = 200,
  seed = 123,
  maxit_M2 = 200,
  maxit_M3 = 200,
  results_file = "bootstrap_LR_M2Iplus_M3Iplus_test.csv"
)

boot_LR_Iplus_test

boot_LR_Iplus <- bootstrap_LR_M2Iplus_vs_M3Iplus(
  fit_M2_I_plus = fit_M2_I_plus,
  fit_M3_I_plus = fit_M3_I_plus,
  x_obs = Ca2,
  h = hCa,
  B = 100,
  h_sim = 0.001,
  burn_obs = 200,
  seed = 123,
  maxit_M2 = 300,
  maxit_M3 = 300,
  results_file = "bootstrap_LR_M2Iplus_M3Iplus.csv"
)

boot_LR_Iplus <- read.csv("bootstrap_LR_M2Iplus_M3Iplus.csv")

successful_LR_Iplus <- boot_LR_Iplus[
  is.finite(boot_LR_Iplus$LR) &
    boot_LR_Iplus$conv_M2 == 0 &
    boot_LR_Iplus$conv_M3 == 0,
]

nrow(successful_LR_Iplus)

LR_obs_Iplus <- 2 * (fit_M2_I_plus$value - fit_M3_I_plus$value)

p_value_LR_Iplus <- (sum(successful_LR_Iplus$LR >= LR_obs_Iplus) + 1) /
  (nrow(successful_LR_Iplus) + 1)

LR_summary_Iplus <- data.frame(
  LR_obs = LR_obs_Iplus,
  p_value = p_value_LR_Iplus,
  n_boot_successful = nrow(successful_LR_Iplus),
  LR_boot_mean = mean(successful_LR_Iplus$LR),
  LR_boot_95 = quantile(successful_LR_Iplus$LR, 0.95),
  row.names = NULL
)

LR_summary_Iplus
```


#And now we do some plots to evaluate the model 3 fitting to the real data
#THESE PLOTS HAVE THE PREDICTION INTERVALS
```{r}
par_hat_M3 <- fit_M3_I_plus$par
expand_M3_par <- function(par) {
  c(
    eta   = unname(par["eta"]),
    a     = unname(par["a"]),
    b     = unname(par["b"]),
    c     = unname(par["c"]),
    d     = unname(par["d"]),
    alpha = unname(par["alpha"]),
    beta  = unname(par["beta"]),
    gamma = unname(par["gamma"])
  )
}

simulate_fine <- function(
    par,
    N_fine,
    h_sim,
    start = c(0, 0)
) {
  par <- expand_M3_par(par)
  
  eta   <- par["eta"]
  a     <- par["a"]
  b     <- par["b"]
  c     <- par["c"]
  d     <- par["d"]
  alpha <- par["alpha"]
  beta  <- par["beta"]
  gamma <- par["gamma"]
  
  X <- numeric(N_fine)
  V <- numeric(N_fine)
  
  X[1] <- start[1]
  V[1] <- start[2]
  
  for (i in 2:N_fine) {
    x_prev <- X[i - 1]
    v_prev <- V[i - 1]
    
    drift_v <- -eta * v_prev +
      a * x_prev^3 +
      b * x_prev^2 +
      c * x_prev +
      d
    
    diff2 <- alpha * v_prev^2 + beta * v_prev + gamma
    diff2 <- max(diff2, 1e-12)
    
    X[i] <- x_prev + h_sim * v_prev
    
    V[i] <- v_prev +
      h_sim * drift_v +
      sqrt(diff2) * sqrt(h_sim) * rnorm(1)
  }
  
  data.frame(X = X, V = V)
}

simulate_student_kramers_observed <- function(
    par,
    N_obs,
    h_obs,
    h_sim = 0.001,
    burn_obs = 200,
    start = c(0, 0)
) {
  ratio <- h_obs / h_sim
  skip <- round(ratio)
  
  N_total_obs <- N_obs + burn_obs + 1
  N_fine <- (N_total_obs - 1) * skip + 1
  
  sim_fine <- simulate_fine(
    par = par,
    N_fine = N_fine,
    h_sim = h_sim,
    start = start
  )
  
  idx <- seq(1, N_fine, by = skip)
  X_obs_all <- sim_fine$X[idx]
  
  X_use <- X_obs_all[(burn_obs + 1):(burn_obs + N_obs + 1)]
  
  X_plot <- X_use[1:N_obs]
  V_plot <- diff(X_use) / h_obs
  
  data.frame(
    X = X_plot,
    V = V_plot
  )
}

compute_density_band <- function(sim_list, component = c("X", "V"), grid) {
  component <- match.arg(component)
  
  dens_mat <- matrix(NA_real_, nrow = length(sim_list), ncol = length(grid))
  
  for (i in seq_along(sim_list)) {
    dd <- density(
      sim_list[[i]][[component]],
      from = min(grid),
      to = max(grid),
      n = length(grid)
    )
    
    dens_mat[i, ] <- dd$y
  }
  
  data.frame(
    grid = grid,
    q025 = apply(dens_mat, 2, quantile, probs = 0.025, na.rm = TRUE),
    q50  = apply(dens_mat, 2, quantile, probs = 0.50,  na.rm = TRUE),
    q975 = apply(dens_mat, 2, quantile, probs = 0.975, na.rm = TRUE)
  )
}

plot_M3_fit <- function(
    par_hat,
    data,
    h_obs = 0.02,
    h_sim = 0.001,
    nsim_band = 200,
    burn_obs = 200,
    seed = 123
) {
  set.seed(seed)
  
  par_hat <- expand_M3_par(par_hat)
  
  age_obs <- data$age
  x_obs <- data$X
  v_obs <- data$V
  
  N_obs <- length(x_obs)
  
  sim_one <- simulate_student_kramers_observed(
    par = par_hat,
    N_obs = N_obs,
    h_obs = h_obs,
    h_sim = h_sim,
    burn_obs = burn_obs,
    start = c(x_obs[1], v_obs[1])
  )
  
  sim_list <- vector("list", nsim_band)
  
  for (i in seq_len(nsim_band)) {
    sim_list[[i]] <- simulate_student_kramers_observed(
      par = par_hat,
      N_obs = N_obs,
      h_obs = h_obs,
      h_sim = h_sim,
      burn_obs = burn_obs,
      start = c(x_obs[1], v_obs[1])
    )
  }
  
  x_lim <- range(x_obs, na.rm = TRUE)
  v_lim <- range(v_obs, na.rm = TRUE)
  
  x_pad <- 0.10 * diff(x_lim)
  v_pad <- 0.10 * diff(v_lim)
  
  x_grid <- seq(
    x_lim[1] - x_pad,
    x_lim[2] + x_pad,
    length.out = 300
  )
  
  v_grid <- seq(
    v_lim[1] - v_pad,
    v_lim[2] + v_pad,
    length.out = 300
  )

  band_x <- compute_density_band(sim_list, component = "X", grid = x_grid)
  band_v <- compute_density_band(sim_list, component = "V", grid = v_grid)
  
  emp_x <- density(x_obs, from = min(x_grid), to = max(x_grid), n = length(x_grid))
  emp_v <- density(v_obs, from = min(v_grid), to = max(v_grid), n = length(v_grid))
  
  sim_x <- density(sim_one$X, from = min(x_grid), to = max(x_grid), n = length(x_grid))
  sim_v <- density(sim_one$V, from = min(v_grid), to = max(v_grid), n = length(v_grid))
  
  df_traj <- data.frame(
    age = age_obs,
    X_obs = x_obs,
    X_sim = sim_one$X,
    V_obs = v_obs,
    V_sim = sim_one$V
  )
  
  p1 <- ggplot(df_traj, aes(x = age)) +
    geom_line(aes(y = X_obs), color = "black", linewidth = 0.45) +
    geom_line(aes(y = X_sim), color = "firebrick", linewidth = 0.4) +
    scale_x_reverse() +
    labs(
      x = NULL,
      y = expression(X[k] == -log(Ca^{2*'+'})),
      title = "Student-Kramers oscillator fitted on Ice core data from Greenland"
    ) +
    theme_minimal(base_size = 12) +
    theme(
      plot.title = element_text(hjust = 0.5, face = "bold"),
      axis.text.x = element_blank()
    )
  
  p2 <- ggplot(df_traj, aes(x = age)) +
    geom_line(aes(y = V_obs), color = "black", linewidth = 0.45) +
    geom_line(aes(y = V_sim), color = "firebrick", linewidth = 0.4) +
    scale_x_reverse() +
    labs(
      x = "Time before 2000 AD [ka]",
      y = expression((X[k+1] - X[k])/h)
    ) +
    theme_minimal(base_size = 12)
  
  p3 <- ggplot() +
    geom_histogram(
      data = data.frame(z = x_obs),
      aes(x = z, y = after_stat(density)),
      bins = 35,
      fill = "grey40",
      color = "grey40",
      alpha = 0.8
    ) +
    geom_ribbon(
      data = band_x,
      aes(x = grid, ymin = q025, ymax = q975),
      fill = "indianred1",
      alpha = 0.45
    ) +
    geom_line(
      data = band_x,
      aes(x = grid, y = q50),
      color = "darkred",
      linewidth = 0.9
    ) +
    geom_line(
      data = data.frame(x = emp_x$x, y = emp_x$y),
      aes(x = x, y = y),
      color = "black",
      linewidth = 0.9
    ) +
    coord_flip() +
    labs(x = NULL, y = NULL) +
    theme_minimal(base_size = 12) +
    theme(axis.text.y = element_blank())
  
  p4 <- ggplot() +
    geom_histogram(
      data = data.frame(z = v_obs),
      aes(x = z, y = after_stat(density)),
      bins = 35,
      fill = "grey40",
      color = "grey40",
      alpha = 0.8
    ) +
    geom_ribbon(
      data = band_v,
      aes(x = grid, ymin = q025, ymax = q975),
      fill = "indianred1",
      alpha = 0.45
    ) +
    geom_line(
      data = band_v,
      aes(x = grid, y = q50),
      color = "darkred",
      linewidth = 0.9
    ) +
    geom_line(
      data = data.frame(x = emp_v$x, y = emp_v$y),
      aes(x = x, y = y),
      color = "black",
      linewidth = 0.9
    ) +
    coord_flip() +
    labs(x = "Density", y = NULL) +
    theme_minimal(base_size = 12) +
    theme(axis.text.y = element_blank())
  
  fig <- (p1 | p3) / (p2 | p4) +
    plot_layout(widths = c(4, 1.25), heights = c(1, 1))
  
  list(
    figure = fig,
    sim_one = sim_one,
    band_x = band_x,
    band_v = band_v
  )
}

out_M3_plot <- plot_M3_fit(
  par_hat = par_hat_M3,
  data = Ca2_final,
  h_obs = hCa,
  h_sim = 0.001,
  nsim_band = 1000,
  burn_obs = 200,
  seed = 123
)

out_M3_plot$figure
```

#One can do the plots for the other models to check if there is one with better 
#fitting to the data... REMOVE THE PREDICTION BANDS TO BE FASTER


#Here is the code used for a different approach in the bx choice.
#(if x_t > 0 then bx+, else bx-)
#This matches the reference paper
#The results here are consistent with what is written there
#the confidence intervals here for b and d include 0

#Required functions and objective function
```{r}
if (!exists("compute_bx")) {
  compute_bx <- function(theta, x_data, branch = "+") {
    a  <- theta["a"]
    b  <- theta["b"]
    Ex <- mean(x_data)
    Vx <- var(x_data)
    disc <- Vx + (Ex + b / (3 * a))^2
    if (branch == "+") -b / (3 * a) + sqrt(disc)
    else               -b / (3 * a) - sqrt(disc)
  }
}

if (!exists("logdet_Dv")) {
  logdet_Dv <- function(state, t, theta, lambda, bx) 0
}

make_A_matrix_both_branches <- function(theta, x_data) {
  eta <- theta["eta"]
  a   <- theta["a"]
  b   <- theta["b"]
  c   <- theta["c"]

  Ex  <- mean(x_data)
  Ex2 <- mean(x_data^2)

  lambda   <- 3 * a * Ex2 + 2 * b * Ex + c
  bx_plus  <- compute_bx(theta, x_data, branch = "+")
  bx_minus <- compute_bx(theta, x_data, branch = "-")

  if (!is.finite(lambda) || !is.finite(bx_plus) || !is.finite(bx_minus)) {
    return(NULL)
  }

  A <- matrix(c(0, 1, lambda, -eta), nrow = 2, byrow = TRUE)

  list(
    A           = A,
    lambda      = lambda,
    bx_plus     = bx_plus,
    bx_minus    = bx_minus,
    b_vec_plus  = c(bx_plus,  0),
    b_vec_minus = c(bx_minus, 0)
  )
}

prepare_omega_I_dynamic <- function(A, diff_par, h) {
  tilde  <- make_par_tilde(diff_par)
  I_mats <- compute_I_matrices(
    A           = A,
    alpha_tilde = tilde$alpha_tilde,
    beta_tilde  = tilde$beta_tilde,
    h           = h
  )

  list(
    I        = I_mats,
    eAh      = expm(A * h),
    diff_par = diff_par
  )
}

omega_I_dynamic_eval <- function(pre, x0, b_vec) {
  I    <- pre$I
  xb   <- as.numeric(x0 - b_vec)

  # Sigma evaluated at b_vec (b_vec[2] = 0, so q22 = gamma)
  Sigma_b <- Cap_sigma(b_vec, pre$diff_par)

  vecOmega <-
    I$I1 %*% as.vector(xb %*% t(xb))        +
    I$I2 %*% as.vector(xb %*% t(b_vec))     +
    I$I3 %*% as.vector(b_vec %*% t(xb))     +
    I$I4 %*% xb                               +
    I$I5 %*% as.vector(Sigma_b)

  Omega <- matrix(vecOmega, nrow = 2, ncol = 2)
  0.5 * (Omega + t(Omega))
}

ss_negloglik_dynamic_branch <- function(par, x_obs, h,
                                         invalid      = Inf,
                                         return_terms = FALSE) {
  theta    <- par[c("eta", "a", "b", "c", "d")]
  diff_par <- par[c("alpha", "beta", "gamma")]

  N <- length(x_obs)

  Y <- cbind(
    x = x_obs[1:(N - 1)],
    v = (x_obs[2:N] - x_obs[1:(N - 1)]) / h
  )

  if (is_bad(Y)) return(invalid)

  x_data <- Y[, "x"]

  lin <- tryCatch(
    make_A_matrix_both_branches(theta, x_data),
    error = function(e) NULL, warning = function(w) NULL
  )
  if (is.null(lin)) return(invalid)

  A      <- lin$A
  lambda <- lin$lambda

  if (is_bad(A)) return(invalid)

  pre_h <- tryCatch(
    prepare_omega_I_dynamic(A, diff_par, h),
    error = function(e) NULL, warning = function(w) NULL
  )
  pre_3h2 <- tryCatch(
    prepare_omega_I_dynamic(A, diff_par, 1.5 * h),
    error = function(e) NULL, warning = function(w) NULL
  )
  if (is.null(pre_h) || is.null(pre_3h2)) return(invalid)

  n_terms <- N - 2
  terms   <- numeric(n_terms)

  for (k in seq_len(n_terms)) {
    z_old <- Y[k, ]
    z_new <- Y[k + 1, ]

    # choose branch by sign of current x
    if (z_old[1] >= 0) {
      bx    <- lin$bx_plus
      b_vec <- lin$b_vec_plus
    } else {
      bx    <- lin$bx_minus
      b_vec <- lin$b_vec_minus
    }

    z_start <- f_flow(z_old, h / 2, theta, lambda, bx)
    if (is_bad(z_start)) return(invalid)

    m <- as.numeric(pre_h$eAh %*% (z_start - b_vec) + b_vec)
    if (is_bad(m)) return(invalid)

    y <- f_flow_inv(z_new, h / 2, theta, lambda, bx)
    if (is_bad(y)) return(invalid)

    r_R <- y[2] - m[2]
    if (!is.finite(r_R)) return(invalid)

    Omega_h <- tryCatch(
      omega_I_dynamic_eval(pre_h,   z_start, b_vec),
      error = function(e) NULL, warning = function(w) NULL
    )
    Omega_3h2 <- tryCatch(
      omega_I_dynamic_eval(pre_3h2, z_start, b_vec),
      error = function(e) NULL, warning = function(w) NULL
    )

    if (is.null(Omega_h) || is.null(Omega_3h2)) return(invalid)
    if (is_bad(Omega_h)  || is_bad(Omega_3h2))  return(invalid)

    Omega_RR_h   <- Omega_h[2, 2]
    Omega_RR_3h2 <- Omega_3h2[2, 2]

    if (!is.finite(Omega_RR_h)   || Omega_RR_h   <= 0) return(invalid)
    if (!is.finite(Omega_RR_3h2) || Omega_RR_3h2 <= 0) return(invalid)

    jac <- logdet_Dv(z_new, h / 2, theta, lambda, bx)
    if (!is.finite(jac)) return(invalid)

    terms[k] <- r_R^2 / Omega_RR_h +
      (2 / 3) * log(Omega_RR_3h2) +
      2 * jac

    if (!is.finite(terms[k])) return(invalid)
  }

  value <- sum(terms)
  if (!is.finite(value)) return(invalid)
  if (return_terms) return(terms)
  value
}

ss_negloglik_M3_dynamic_branch <- function(par, x_obs, h,
                                            invalid      = Inf,
                                            return_terms = FALSE) {
  ss_negloglik_dynamic_branch(par, x_obs, h, invalid, return_terms)
}

objective_M3_dynamic_branch_for_optim <- function(par, x_obs, h) {
  invalid_finite <- 1e100
  val <- tryCatch(
    ss_negloglik_M3_dynamic_branch(par, x_obs, h,
                                    invalid = invalid_finite),
    error = function(e) invalid_finite
  )
  if (!is.finite(val)) invalid_finite else val
}

fit_M3_dynamic <- optim(
  par     = theta_start_M3,
  fn      = objective_M3_dynamic_branch_for_optim,
  x_obs   = Ca2,
  h       = hCa,
  method  = "L-BFGS-B",
  lower   = lower_bounds_M3,
  upper   = upper_bounds_M3,
  control = list(maxit = 1000, factr = 1e7, trace = 1)
)
```

#it is not difficult to do the same for model 2...
#one can use the plots defined above to check the model fitting... A suggestion 
#is that the prediction intervals are removed in order to get faster results...

#one can check now the confidence intervals for b and d with a bootstrap 
#we can also do a parametric bootstrap LR test to see which model to chose between these last 2
#the results will not be consistent with the other approach used.

```{r}
citation("parallel")
```

#goodness of fitness not working properly + some sanity tests of it
```{r}
library(expm)
library(parallel)

is_bad <- function(x) any(!is.finite(x))

make_emp_stats <- function(x_obs, h) {
  x_data <- x_obs[1:(length(x_obs) - 1)]
  list(
    Ex  = mean(x_data),
    Vx  = var(x_data),
    Ex2 = mean(x_data^2)
  )
}

compute_bx <- function(theta, emp_stats, branch = "+") {
  a  <- theta["a"]
  b  <- theta["b"]
  Ex <- emp_stats$Ex
  Vx <- emp_stats$Vx
  disc <- Vx + (Ex + b / (3 * a))^2
  if (branch == "+") -b / (3 * a) + sqrt(disc)
  else               -b / (3 * a) - sqrt(disc)
}

make_A_matrix <- function(theta, emp_stats, branch = "+") {
  eta <- theta["eta"]; a <- theta["a"]; b <- theta["b"]; c <- theta["c"]
  lambda <- 3 * a * emp_stats$Ex2 + 2 * b * emp_stats$Ex + c
  bx     <- compute_bx(theta, emp_stats, branch)
  A      <- matrix(c(0, lambda, 1, -eta), nrow = 2, byrow = TRUE)
  list(A = A, b_vec = c(bx, 0), lambda = lambda, bx = bx)
}

make_par_tilde <- function(diff_par) {
  alpha_tilde <- matrix(0, 4, 4); beta_tilde <- matrix(0, 4, 2)
  alpha_tilde[4, 4] <- diff_par["alpha"]
  beta_tilde[4, 2]  <- diff_par["beta"]
  list(alpha_tilde = alpha_tilde, beta_tilde = beta_tilde,
       gamma_vec = c(0, 0, 0, diff_par["gamma"]))
}

ap_integral <- function(M, N, P, h) {
  r <- nrow(M); cc <- nrow(P)
  B <- rbind(cbind(M, N), cbind(matrix(0, cc, r), P))
  E <- expm(B * h)
  E[1:r, (r + 1):(r + cc)]
}

compute_I_matrices <- function(A, alpha_tilde, beta_tilde, h) {
  K        <- kronecker(diag(2), A) + kronecker(A, diag(2))
  M        <- K + alpha_tilde
  I_kron_A <- kronecker(diag(2), A)
  A_kron_I <- kronecker(A, diag(2))
  list(
    I1 = ap_integral(M, alpha_tilde, K,               h),
    I2 = ap_integral(M, alpha_tilde, I_kron_A,        h),
    I3 = ap_integral(M, alpha_tilde, A_kron_I,        h),
    I4 = ap_integral(M, beta_tilde,  A,               h),
    I5 = ap_integral(M, diag(4),     matrix(0, 4, 4), h)
  )
}

Cap_sigma <- function(b_vec, diff_par) {
  v     <- b_vec[2]
  q22   <- diff_par["alpha"] * v^2 + diff_par["beta"] * v + diff_par["gamma"]
  matrix(c(0, 0, 0, q22), nrow = 2)
}

prepare_omega_I <- function(A, b_vec, diff_par, h) {
  tilde  <- make_par_tilde(diff_par)
  I_mats <- compute_I_matrices(A, tilde$alpha_tilde, tilde$beta_tilde, h)
  list(
    method      = "I",
    I           = I_mats,
    eAh         = expm(A * h),
    b_vec       = b_vec,
    Sigma_b_vec = as.vector(Cap_sigma(b_vec, diff_par))
  )
}

omega_I_eval_vec <- function(pre, z_start_mat) {
  I <- pre$I; b_vec <- pre$b_vec
  n  <- nrow(z_start_mat)
  xb <- sweep(z_start_mat, 2, b_vec, "-")
  
  xb_outer   <- cbind(xb[,1]*xb[,1], xb[,2]*xb[,1], xb[,1]*xb[,2], xb[,2]*xb[,2])
  xb_b_outer <- cbind(xb[,1]*b_vec[1], xb[,2]*b_vec[1], xb[,1]*b_vec[2], xb[,2]*b_vec[2])
  b_xb_outer <- cbind(b_vec[1]*xb[,1], b_vec[1]*xb[,2], b_vec[2]*xb[,1], b_vec[2]*xb[,2])
  
  result <- xb_outer   %*% t(I$I1) +
    xb_b_outer %*% t(I$I2) +
    b_xb_outer %*% t(I$I3) +
    xb         %*% t(I$I4) +
    matrix(pre$Sigma_b_vec, nrow = n, ncol = 4, byrow = TRUE) %*% t(I$I5)
  
  result[, 2] <- result[, 3] <- 0.5 * (result[, 2] + result[, 3])
  result
}


N_fun_vec <- function(states, theta, lambda, bx) {
  x  <- states[, 1]
  n2 <- theta["a"]*x^3 + theta["b"]*x^2 + theta["c"]*x + theta["d"] - lambda*(x - bx)
  cbind(0, n2)
}

f_flow_vec     <- function(states, t, theta, lambda, bx) states + t * N_fun_vec(states, theta, lambda, bx)
f_flow_inv_vec <- function(states, t, theta, lambda, bx) states - t * N_fun_vec(states, theta, lambda, bx)

ss_negloglik_M3_partial_R <- function(
    par,
    x_obs,
    h,
    branch       = "+",
    omega_method = "I",
    invalid      = 1e100,
    return_terms = FALSE,
    emp_stats    = NULL
) {
  theta    <- par[c("eta", "a", "b", "c", "d")]
  diff_par <- par[c("alpha", "beta", "gamma")]
  
  Y <- cbind(x = x_obs[1:(N-1)], v = diff(x_obs) / h)
  
  lin <- make_A_matrix(theta, emp_stats, branch)

  
  A <- lin$A; b_vec <- lin$b_vec; lambda <- lin$lambda; bx <- lin$bx
  
  pre_h   <- prepare_omega_I(A, b_vec, diff_par, h)
  pre_3h2 <- prepare_omega_I(A, b_vec, diff_par, 1.5 * h)
  
  n_terms <- N - 2
  z_old   <- Y[1:n_terms,       , drop = FALSE]
  z_new   <- Y[2:(n_terms+1),   , drop = FALSE]
  
  z_start <- f_flow_vec(z_old, h/2,  theta, lambda, bx)
  y       <- f_flow_inv_vec(z_new, h/2, theta, lambda, bx)
  
  m <- t(pre_h$eAh %*% t(sweep(z_start, 2, b_vec, "-")))
  m <- sweep(m, 2, b_vec, "+")
  
  r_R <- y[, 2] - m[, 2]
  
  Omega_h_mat   <- omega_I_eval_vec(pre_h,   z_start)
  Omega_3h2_mat <- omega_I_eval_vec(pre_3h2, z_start)
  
  Omega_RR_h   <- Omega_h_mat[, 4]
  Omega_RR_3h2 <- Omega_3h2_mat[, 4]
  
  
  terms <- r_R^2 / Omega_RR_h + (2/3) * log(Omega_RR_3h2)
  
  if (return_terms) return(terms)
  sum(terms)
}


#ios below, above is just the likelihood in a faster way


make_emp_stats <- function(x_obs, h) {
  x_data <- x_obs[1:(length(x_obs) - 1)]
  list(Ex = mean(x_data), Vx = var(x_data), Ex2 = mean(x_data^2))
}

get_terms_M3_I_plus <- function(par, x_obs, h, invalid = 1e100, emp_stats = NULL) {
  terms <- tryCatch(
    ss_negloglik_M3_partial_R(par, x_obs, h, "+", "I", invalid, TRUE, emp_stats),
    error = function(e) NULL
  )
  if (is.null(terms)) return(NULL)
  if (length(terms) == 1 && terms >= invalid * 0.99) return(NULL)
  if (any(!is.finite(terms))) return(NULL)
  terms
}

objective_delete_one_M3_I_plus <- function(par, x_obs, h, k_delete,
                                           invalid = 1e100, emp_stats = NULL) {
  terms <- get_terms_M3_I_plus(par, x_obs, h, invalid, emp_stats)
  if (is.null(terms)) return(invalid)
  if (k_delete < 1 || k_delete > length(terms)) return(invalid)
  value <- sum(terms[-k_delete])
  if (!is.finite(value)) return(invalid)
  value
}

fit_delete_one_M3_I_plus <- function(k_delete, fit_full, x_obs, h, lower, upper,
                                     maxit = 2000, factr = 1e6,
                                     invalid = 1e100, emp_stats = NULL) {
  fit <- tryCatch(
    optim(
      par = fit_full$par, fn = objective_delete_one_M3_I_plus,
      x_obs = x_obs, h = h, k_delete = k_delete,
      invalid = invalid, emp_stats = emp_stats,
      method = "L-BFGS-B", lower = lower, upper = upper,
      control = list(maxit = maxit, trace = 0, factr = factr)
    ),
    error = function(e) NULL
  )
  if (is.null(fit) || !is.finite(fit$value) || fit$value >= 0.99 * invalid) return(NULL)
  fit
}

make_local_bounds <- function(par_full, lower_global, upper_global,
                              rel_width = 0.10, abs_width = 1e-4) {
  width <- pmax(abs(par_full) * rel_width, abs_width)
  list(
    lower = pmax(lower_global, par_full - width),
    upper = pmin(upper_global, par_full + width)
  )
}

IOS_subset_M3_I_plus_fast <- function(
    fit_full,
    x_obs,
    h,
    m_subset     = 500,
    lower        = lower_bounds_M3,
    upper        = upper_bounds_M3,
    seed         = 123,
    maxit        = 2000,
    factr        = 1e6,
    results_file = NULL,
    n_cores      = 1,
    r            = 5,
    s            = 3,
    invalid      = 1e100,
    emp_stats    = NULL,
    verbose      = TRUE
) {
  set.seed(seed)
  
  if (is.null(emp_stats)) emp_stats <- make_emp_stats(x_obs, h)
  
  full_terms <- get_terms_M3_I_plus(fit_full$par, x_obs, h, invalid, emp_stats)
  if (is.null(full_terms)) stop("Could not compute full-model likelihood terms.")
  
  N_terms  <- length(full_terms)
  m_subset <- min(m_subset, N_terms)
  k_subset <- sort(sample(seq_len(N_terms), size = m_subset, replace = FALSE))
  
  if (verbose) {
    cat("IOS | N_terms =", N_terms, "| m_subset =", m_subset, "| n_cores =", n_cores, "\n")
  }
  
  one_delete_fit <- function(k) {
    t0 <- proc.time()[["elapsed"]]
    
    fit_loo <- fit_delete_one_M3_I_plus(
      k_delete = k, fit_full = fit_full, x_obs = x_obs, h = h,
      lower = lower, upper = upper, maxit = maxit, factr = factr,
      invalid = invalid, emp_stats = emp_stats
    )
    
    elapsed <- proc.time()[["elapsed"]] - t0
    
    empty <- data.frame(k=k, full_term=NA_real_, loo_term=NA_real_,
                        delta=NA_real_, loo_nll=NA_real_,
                        convergence=NA_integer_, time_sec=elapsed)
    
    if (is.null(fit_loo)) return(empty)
    
    loo_terms <- get_terms_M3_I_plus(fit_loo$par, x_obs, h, invalid, emp_stats)
    if (is.null(loo_terms)) return(empty)
    
    data.frame(
      k           = k,
      full_term   = full_terms[k],
      loo_term    = loo_terms[k],
      delta       = loo_terms[k] - full_terms[k],
      loo_nll     = fit_loo$value,
      convergence = fit_loo$convergence,
      time_sec    = elapsed
    )
  }
  
  if (n_cores <= 1) {
    results_list <- vector("list", length(k_subset))
    for (j in seq_along(k_subset)) {
      if (verbose) cat("  fit", j, "/", m_subset, "| k =", k_subset[j], "\n")
      results_list[[j]] <- one_delete_fit(k_subset[j])
    }
  } else {
    n_cores <- min(n_cores, parallel::detectCores())
    cl <- parallel::makeCluster(n_cores)
    on.exit(try(parallel::stopCluster(cl), silent = TRUE), add = TRUE)
    parallel::clusterExport(cl, varlist = ls(envir = .GlobalEnv), envir = .GlobalEnv)
    parallel::clusterEvalQ(cl, library(expm))
    results_list <- parallel::parLapplyLB(cl, k_subset, one_delete_fit)
  }
  
  results <- do.call(rbind, results_list)
  results <- results[order(results$k), ]
  rownames(results) <- NULL
  
  if (!is.null(results_file)) write.csv(results, results_file, row.names = FALSE)
  
  successful <- results[is.finite(results$delta) &
                          !is.na(results$convergence) &
                          results$convergence == 0, ]
  
  n_success <- nrow(successful)
  if (n_success == 0) stop("No successful delete-one fits.")
  
  T_unscaled <- sum(successful$delta)
  T_scaled   <- (N_terms / n_success) * T_unscaled
  ref        <- r + (9/4) * s
  
  if (verbose) {
    cat("IOS done | n_success =", n_success, "| T_scaled =",
        round(T_scaled, 3), "| ref =", ref, "\n")
  }
  
  list(
    model = "M3", branch = "+", omega_method = "I",
    N_terms = N_terms, m_subset_requested = m_subset,
    n_success = n_success,
    T_subset_unscaled = T_unscaled,
    T_subset_scaled   = T_scaled,
    reference_value   = ref,
    emp_stats = emp_stats,
    results   = results,
    successful = successful
  )
}

simulate_M3 <- function(N_obs, h_obs, h_sim, par, start = c(0, 0)) {
  theta    <- par[c("eta", "a", "b", "c", "d")]
  diff_par <- par[c("alpha", "beta", "gamma")]
  
  skip   <- round(h_obs / h_sim)
  N_fine <- (N_obs - 1) * skip + 1
  
  X <- numeric(N_fine); V <- numeric(N_fine)
  X[1] <- start[1];     V[1] <- start[2]
  
  for (i in 2:N_fine) {
    xp <- X[i-1]; vp <- V[i-1]
    drift_v <- -theta["eta"]*vp + theta["a"]*xp^3 +
      theta["b"]*xp^2 + theta["c"]*xp + theta["d"]
    sigma   <- sqrt(pmax(diff_par["alpha"]*vp^2 +
                           diff_par["beta"]*vp +
                           diff_par["gamma"], 1e-10))
    X[i] <- xp + h_sim * vp
    V[i] <- vp + h_sim * drift_v + sigma * sqrt(h_sim) * rnorm(1)
  }
  
  X[seq(1, N_fine, by = skip)]
}

bootstrap_IOS_M3 <- function(
    fit_observed,       # the fitted model on your real data
    x_obs,              # your real data
    h,                  # observation step
    h_sim      = 0.001,
    B          = 199,
    m_subset   = 200,   # smaller than 500 for speed
    maxit      = 1000,
    factr      = 1e7,
    rel_width  = 0.10,
    lower_global = lower_bounds_M3,
    upper_global = upper_bounds_M3,
    r          = 5,
    s          = 3,
    n_cores    = 4,     # parallelize over bootstrap replicates
    seed       = 42
) {
  set.seed(seed)
  
  par_true <- fit_observed$par
  N_obs    <- length(x_obs)
  ref      <- r + (9/4) * s
  
  # Worker function: one bootstrap replicate
  one_bootstrap <- function(b) {
    set.seed(seed + b)
    
    # 1. Simulate
    x_b <- tryCatch(
      simulate_M3(N_obs, h, h_sim, par_true, start = c(x_obs[1], 0)),
      error = function(e) NULL
    )
    if (is.null(x_b) || any(!is.finite(x_b))) return(NA_real_)
    
    # 2. Freeze emp stats for this bootstrap dataset
    emp_b <- make_emp_stats(x_b, h)
    
    # 3. Refit
    fit_b <- tryCatch(
      optim(
        par = par_true,
        fn  = function(par) {
          v <- ss_negloglik_M3_partial_R(par, x_b, h, "+", "I",
                                         1e100, FALSE, emp_b)
          if (!is.finite(v)) 1e100 else v
        },
        method  = "L-BFGS-B",
        lower   = lower_global,
        upper   = upper_global,
        control = list(maxit = maxit, factr = factr, trace = 0)
      ),
      error = function(e) NULL
    )
    
    if (is.null(fit_b) || fit_b$convergence != 0) return(NA_real_)
    
    # 4. Local bounds around bootstrap MLE
    lb <- make_local_bounds(fit_b$par, lower_global, upper_global, rel_width)
    
    # 5. IOS on bootstrap dataset
    ios_b <- tryCatch(
      IOS_subset_M3_I_plus_fast(
        fit_full  = fit_b,
        x_obs     = x_b,
        h         = h,
        m_subset  = m_subset,
        lower     = lb$lower,
        upper     = lb$upper,
        maxit     = maxit,
        factr     = factr,
        n_cores   = 1,        # already parallel over b
        emp_stats = emp_b,
        verbose   = FALSE
      ),
      error = function(e) NULL
    )
    
    if (is.null(ios_b)) return(NA_real_)
    ios_b$T_subset_scaled
  }
  
  cat("Running", B, "bootstrap replicates on", n_cores, "cores...\n")
  
  if (n_cores <= 1) {
    TN_boot <- sapply(seq_len(B), function(b) {
      cat("Bootstrap", b, "/", B, "\n")
      one_bootstrap(b)
    })
  } else {
    cl <- parallel::makeCluster(n_cores)
    on.exit(try(parallel::stopCluster(cl), silent = TRUE), add = TRUE)
    parallel::clusterExport(cl, varlist = ls(envir = .GlobalEnv), envir = .GlobalEnv)
    parallel::clusterEvalQ(cl, library(expm))
    TN_boot <- parallel::parSapply(cl, seq_len(B), one_bootstrap)
  }
  
  TN_clean <- TN_boot[is.finite(TN_boot)]
  n_ok     <- length(TN_clean)
  cat("Successful replicates:", n_ok, "/", B, "\n")
  
  T_obs   <- fit_observed$T_subset_scaled  # pass the real IOS result in
  p_value <- mean(TN_clean >= T_obs)
  crit_95 <- quantile(TN_clean, 0.95)
  crit_99 <- quantile(TN_clean, 0.99)
  
  cat("\n--- Bootstrap results ---\n")
  cat("Observed T_scaled:      ", round(T_obs,   3), "\n")
  cat("Bootstrap mean:         ", round(mean(TN_clean), 3), "\n")
  cat("Asymptotic reference:   ", ref, "\n")
  cat("Bootstrap 95th pct:     ", round(crit_95, 3), "\n")
  cat("Bootstrap 99th pct:     ", round(crit_99, 3), "\n")
  cat("Bootstrap p-value:      ", round(p_value, 3), "\n")
  
  hist(TN_clean, breaks = 30,
       main = "Bootstrap null distribution of T_N",
       xlab = "T_N")
  abline(v = T_obs,   col = "red",  lwd = 2)
  abline(v = crit_95, col = "blue", lwd = 2, lty = 2)
  legend("topright",
         legend = c(paste0("Observed (", round(T_obs,2), ")"),
                    paste0("95th pct (", round(crit_95,2), ")")),
         col = c("red", "blue"), lty = c(1, 2), lwd = 2)
  
  list(
    T_observed     = T_obs,
    TN_boot        = TN_boot,
    TN_clean       = TN_clean,
    n_success      = n_ok,
    p_value        = p_value,
    crit_95        = crit_95,
    crit_99        = crit_99,
    reference_asymptotic = ref
  )
}

emp_stats_M3 <- make_emp_stats(Ca2, hCa)

fit_M3_I_plus_fixed_full <- optim(
  par = fit_M3_I_plus$par,
  fn  = function(par) {
    v <- ss_negloglik_M3_partial_R(par, Ca2, hCa, "+", "I", 1e100, FALSE, emp_stats_M3)
    if (!is.finite(v)) 1e100 else v
  },
  method  = "L-BFGS-B",
  lower   = lower_bounds_M3,
  upper   = upper_bounds_M3,
  control = list(maxit = 2000, factr = 1e7, trace = 1)
)
cat("Convergence:", fit_M3_I_plus_fixed_full$convergence, "\n")
cat("Value:      ", fit_M3_I_plus_fixed_full$value, "\n")


local_bounds_10 <- make_local_bounds(
  fit_M3_I_plus_fixed_full$par, lower_bounds_M3, upper_bounds_M3, 0.10
)

IOS_M3_I_plus_final <- IOS_subset_M3_I_plus_fast(
  fit_full     = fit_M3_I_plus_fixed_full,
  x_obs        = Ca2,
  h            = hCa,
  m_subset     = 500,
  lower        = local_bounds_10$lower,
  upper        = local_bounds_10$upper,
  seed         = 123,
  maxit        = 2000,
  factr        = 1e6,
  results_file = "IOS_M3_final.csv",
  n_cores      = 4,
  emp_stats    = emp_stats_M3
)


fit_M3_I_plus_fixed_full$T_subset_scaled <- IOS_M3_I_plus_final$T_subset_scaled

boot_results <- bootstrap_IOS_M3(
  fit_observed = fit_M3_I_plus_fixed_full,
  x_obs        = Ca2,
  h            = hCa,
  h_sim        = 0.001,
  B            = 199,
  m_subset     = 200,
  maxit        = 1000,
  factr        = 1e7,
  rel_width    = 0.10,
  lower_global = lower_bounds_M3,
  upper_global = upper_bounds_M3,
  n_cores      = 4,
  seed         = 42
)

results_fixed <- IOS_M3_I_plus_final$results

results_fixed$delta <- results_fixed$full_term - results_fixed$loo_term

successful_fixed <- results_fixed[
  is.finite(results_fixed$delta) &
    !is.na(results_fixed$convergence) &
    results_fixed$convergence == 0, ]

T_scaled_correct <- (IOS_M3_I_plus_final$N_terms / nrow(successful_fixed)) *
  sum(successful_fixed$delta)

#other result (fliepped) which one is correct?
TN_boot_correct <- -boot_results$TN_boot   # just flip signs

TN_clean_correct <- TN_boot_correct[is.finite(TN_boot_correct)]
p_value_correct  <- mean(TN_clean_correct >= T_scaled_correct)
crit_95_correct  <- quantile(TN_clean_correct, 0.95)
crit_99_correct <- quantile(TN_clean_correct, 0.99)

cat("Corrected T_scaled:   ", round(T_scaled_correct, 3), "\n")
cat("Corrected 95th pct:   ", round(crit_95_correct,  3), "\n")
cat("Corrected 99th pct:   ", round(crit_99_correct,  3), "\n")
cat("Corrected p-value:    ", round(p_value_correct,  3), "\n")
cat("Reference (asymptotic):", IOS_M3_I_plus_final$reference_value, "\n") 
```

